---
title: Distribution approximation via quantiles
date: "`r format(Sys.time(), '%H:%M %d %B %Y')`"
---

Problem statement: given a small number of quantiles (3-5?), find a good approximation to a univariate distribution that can be used for various tasks (e.g. generating random deviates for Bayesian calculations)

Start with a baseline distribution with an appropriate domain. For example, for a distribution that has support on the whole real line ($(-\infty,\infty)$), we might use the Normal distribution or, more conveniently (see below), the logistic distribution. Call the cumulative distribution function (CDF) of this distribution $g^{-1}$: in other words, we will treat the CDF as an *inverse link function* in the sense of generalized models, and its inverse $g$ as a link function. For the logistic distribution, $g$ is the logit function $\log(x/(1-x))$.

Now, given a vector of quantiles $q$ and a vector of values $x$,

- set up a graph $\{g(x),q\}$ on the scale of the CDF.
- fit a **monotonic** interpolation spline $S(g(x))$ through the values $\{g(x),q\}$, augmented by the anchoring points at (0,0) and (1,1) defining the domain of the distribution.  This is the estimate of the (link-scale) CDF we want.

Now we can get:

- random deviates: pick uniform deviates $u$ and transform them via $g^{-1}(S^{-1}(u)
- cumulative distribution function on the original scale: $S(x)$ transformed appropriately (multiply/divide by the derivative of the (inverse?) link function)
- density function on the original scale: $S'(x)$ transformed appropriately

What's nice about this approach is that the null case (*no* quantiles given) reverts to the specified baseline distribution, while additional information will allow us to deviate from the baseline.

The only concerns I have about this approach are (1) it won't give any kind of interpretable analytic form (although I didn't think that was high on the list of desiderata); (2) I'm not sure what the implications are about tail behavior; (3) I don't really have much sense about the mathematical/statistical properties of a function that is piecewise cubic on the link scale.

## Example

Use logistic distribution as a baseline, specify that the median is at 0.2 (rather than 0):

```{r fit1}
xvec <- 0.2
lxvec <- plogis(xvec) ## apply link function
qvec <- 0.5
q_aug <- c(0,qvec,1)
x_aug <- c(0,lxvec,1)
par(las=1,bty="l",xaxs="i",yaxs="i")
plot(x_aug,q_aug,xlab="g(x)",ylab="quantile",pch=16,xpd=NA)
abline(a=0,b=1)
library(splines)
## save values
xx <- xspline(x_aug,q_aug,shape=-1,draw=FALSE)
lines(xx)
if (FALSE) {
    ## fails: minimum number of points?
    ii <- interpSpline(x_aug,q_aug)
}
```

Now we just need to be able to invert the results of xspline ...
Crudely:

```{r plot1}
par(las=1,bty="l")
xspline_inv <- approxfun(xx$y,xx$x)
ry <- qlogis(xspline_inv(runif(10000)))
hist(ry,breaks=100,col="gray",prob=TRUE)
curve(dlogis(x),add=TRUE,col=2,lwd=2)
```

Try more quantiles:
```{r fit2}
xvec <- c(0,0.2,0.5)
lxvec <- plogis(xvec) ## apply link function
qvec <- c(0.25,0.5,0.75)
q_aug <- c(0,qvec,1)
x_aug <- c(0,lxvec,1)
par(las=1,bty="l",xaxs="i",yaxs="i")
plot(x_aug,q_aug,xlab="g(x)",ylab="quantile",pch=16,xpd=NA)
abline(a=0,b=1)
library(splines)
## tweak shape parameter until CDF is monotonic ...
xx <- xspline(x_aug,q_aug,shape=-0.3,draw=FALSE)
lines(xx,col=4)
all(diff(xx$x)>=0)  ##  OK
ii <- interpSpline(x_aug,q_aug)
x_aug2 <- seq(0,1,length=101)
lines(predict( ii, x_aug2), col=2)
```

```{r plot2}
par(las=1,bty="l")
xspline_inv <- approxfun(xx$y,xx$x)
ry <- qlogis(xspline_inv(runif(10000)))
hist(ry,breaks=100,col="gray",prob=TRUE)
curve(dlogis(x),add=TRUE,col=2,lwd=2)
```
