---
title: "derivatives (especially higher-order) in R"
---

Inspired by [this tweet](https://twitter.com/BjornstadOttar/status/1535773594252353537): what are our options for calculating derivatives, and especially higher-order derivatives, in R?

```{r}
remotes::install_github("quantumelixir/radx")
```

```{r pkgs}
library(Ryacas)
library(madness)
library(cOde)
library(calculus)
library(numDeriv)
library(radx)
```

- are functions entered as expressions, character strings, or functions?
- results as expressions, char strings, or functions?
- numeric, symbolic, or automatic?


```{r king_bjornstad, echo=FALSE}
## Bjørnstad/King function
KB_Jacobian <- function (.vars, ...) {
    ## capture the elements
    vf <- substitute(list(...))[-1L]
    ## capture
    vars <- sapply(substitute(.vars), deparse)
    if (length(vars)>1) vars <- vars[-1L]
    ## differentiatiate each of the components
    sapply(
        vars,
        \(var) sapply(vf,D,name=var)
   ) -> jac
   ## set up dimensions for the output
   dd <- c(length(vf),length(vars))
   dim(jac) <- NULL
   ## figure out names
   dn <- list(
      ifelse(
        nzchar(names(vf)),
        names(vf),
        sapply(vf,deparse)
    ),
    vars
   )
  ## the hard part
  fun <- function (...) {
    ## figure out the correct environment in which to evaluate everything  
    e <- c(as.list(sys.frame(sys.nframe())),...)
    ## evaluate each element of the jacobian
    J <- vapply(jac,eval,numeric(1L),envir=e)
    ## reset the dimensions and names  
    dim(J) <- dd
    dimnames(J) <- dn
    J
  }
  ## !!
  formals(fun) <- eval(
    parse(text=paste0("alist(",paste0(c(vars,"..."),"=",collapse=","),")"))
  )
  ## return function as the result
  fun
}
```

```{r}
f_kb <- KB_Jacobian(.vars = c(x,y), sin(x), cos(x), atan(y/x), tan(x+y))
f_kb(y=2,x=3)
```

## cOde

```{r}
## utility for "lifting" a character vector/matrix to a function that returns numeric values
evalwrap <- function(c) {
    function(...) {
        ee <- list2env(list(...))
        if (is.matrix(c)) {
            nc <- ncol(c); nr <- nrow(c)
        } else {
            nc <- length(ls(ee)) ## number of vars
            nr <- length(c)/nc
        }
        m <- matrix(NA_real_, ncol = nc, nrow = nr)
        m[] <- vapply(c,
                      function(x) eval(parse(text = x), envir = ee),
                      FUN.VALUE = numeric(1))
        return(m)
    }
}
```

```{r}
library(cOde)
charvec <- c(f1 = "sin(x)", f2 = "cos(x)", f3 = "atan(y/x)", f4 = "tan(x+y)")
j_code <- jacobianSymb(f = charvec, variables = c("x", "y"))
f_code <- evalwrap(j_code)
f_code(y=2,x=3)
```

- input as character, output as character (would need to `eval(parse(...))`)
- input returned as unpacked vector (would have to re-pack into a matrix)
- function components must be named
- uses `deparse(D(...))` internally

```{r}
library(calculus)
j_calc <- calculus::jacobian(charvec, var = c("x", "y"))
f_calc <- evalwrap(j_calc)
f_calc(y=2,x=3)
```

- input as character, output as character (would need to `eval(parse(...))`)
- input returned as matrix (dimensions not named)


## also: numDeriv, Deriv, madness, TMB (pytorch thing), (Nash autodiff thing?)

## symbolic and numerical?
## user-extended derivs table?
## autodiff?


```
