### combinations of thin-plate splines and phylogenetic effects

### separable

`Z = t(rt$Zt)` orders its columns `[species-outer, trait-inner]` (each species' intercept/slope pair adjacent), so `b` (an `ntip x 2` matrix) must be flattened to match via `c(t(b))`, not `c(b)`

```{r fit_sep1}
p2 <- list(beta = rep(0, 2),
           b = matrix(0,
                      nrow = nrow(vcmat),
                      ncol = 2),
           logrsd=rep(0, 2),
           logsdres = 0,
           corval=0)
chdat_x <- c(chdat, lst(X, phylomat = vcmat, scale = 1, Z = t(rt$Zt), us2))
obj2 <- MakeADFun(nllfun_sep, p2, silent = TRUE, random = "b")
fit_sep <- TMBfit(obj2)
```

As a ground-truth check, we can also write out the *dense* Kronecker covariance directly: stack the 49 species x 2 traits into one 98-vector `b` with `b ~ N(0, Sigma2 %x% vcmat)`, no `dseparable()` abstraction involved.

```{r dense_slopes}
Zdense <- t(rt$Zt)
p0_ds <- modifyList(p0, list(b = rep(0, 2*nrow(vcmat)), logrsd = rep(0,2), corval = 0))
chdat_x <- c(chdat, lst(X, Zdense, vcmat, us2))
obj_ds <- MakeADFun(nllfun_dense_slopes, p0_ds, silent = TRUE, random = "b")
fit_dense_slopes <- TMBfit(obj_ds)
```

`fixef(fit_sep)` and `fixef(fit_dense_slopes)` now agree (as do their `logLik`s), confirming `dseparable()` really does implement the `Sigma2 %x% vcmat` covariance.

Now the edge-based version, answering the to-do question below ("does the Z-matrix random-slopes model match the separable model?"): `KhatriRao(t(pZ), t(J))` builds a `nobs x (2*nedge)` matrix whose columns are ordered `[edge-outer, trait-inner]` -- each edge contributes an (intercept-innovation, slope-innovation) pair. Propagating an edge-level, iid-across-edges, 2x2-correlated Brownian innovation up the tree induces `Cov(trait_a_i, trait_b_j) = Sigma2[a,b] * vcmat[i,j]` for species `i,j` -- exactly the same separable covariance as above, just built without ever forming `vcmat` or a covariance/precision matrix at all.

```{r phylo_edge_slopes}
f <- ~ 1 + (1 + log_bm | species)
J <- eval(bquote(model.matrix(~.(findbars(f)[[1]][[2]]), data = chdat)))
pZ <- phylo_to_Z(chtree)
## reorder pZ's tip-label rows to match chdat's (species-factor) row order
pZ_ord <- pZ[as.character(chdat$species), ]
KR <- t(KhatriRao(t(pZ_ord), t(J)))

p0_es <- modifyList(p0, list(b = rep(0, 2*nrow(chtree$edge)), logrsd = rep(0,2), corval = 0))
chdat_x <- c(chdat, lst(X, KR, us2))
obj_es <- MakeADFun(nllfun_edge_slopes, p0_es, silent = TRUE, random = "b")
fit_edge_slopes <- TMBfit(obj_es)
```

`fixef(fit_edge_slopes)` and `logLik(fit_edge_slopes)` match `fit_sep`/`fit_dense_slopes` too (all three agree to several significant figures) -- so yes, the edge-based `Z`-matrix random-slopes model is exactly the separable model, just reparameterized.

### tensor-product

```{r tp_tests}
## FIXME: tensor.prod.model.matrix()/tensor.prod.penalties() both use the
## convention "first list element = outer block, second = inner" (verified
## against small test matrices in phyloslopes_tiny.R/phyloslopes_tiny_explore.R). tX below is built from
## list(Z, X) (phylo outer, trait inner) but tp is built from list(diag(2),
## solve(vcmat)) (trait outer, phylo inner) -- reversed, so tp's and tX's
## columns don't actually line up. Harmless for now since tp/tX are never
## combined into a fitted nllfun_* here, but fix the order (phylo first) if
## this ever gets wired into a real fit.
tp <- tensor.prod.penalties(list(diag(2), solve(vcmat)))
lapply(tp, dim)
plot_grid(plotlist = lapply(tp, ifun))
tX <- tensor.prod.model.matrix(list(Z, as(X, "dgCMatrix")))
dim(tX)
```

**Is the separable random-slopes model (above) nested in this tensor-product model?** No -- see `README_tensor.qmd` for the Kronecker-product-vs-Kronecker-sum non-nesting proof, the saddle-point fit that results from trying anyway, and the follow-up `diag(2)`-specific non-identifiability finding.

## phylogenetic splines

McGillycuddy *et al.*'s `propto` manuscript (`main.tex`) fits an *additive* model combining a penalized thin-plate-spline smoother of `log_bm` with a phylogenetic random intercept:
```
glmmTMB(log(runningspeed) ~ s(logbm) + propto(0 + species | g, mat), data = data,
        control = glmmTMBControl(optimizer = optim, optArgs = list(method = "BFGS")))
```
the two random effects (spline wiggle, phylogenetic intercept) are completely independent, summed in the linear predictor. Here we redo this comparing three structures: (1) that additive model, (2) a *separable* model where the spline's wiggly coefficients also get their own (single-scale) phylogenetically-correlated component, and (3) a genuine tensor-product smooth of phylogeny x `log_bm`.

```{r spline_setup}
## thin-plate spline marginal for log_bm; absorb.cons=TRUE matches glmmTMB's
## internal s() handling (its own sum-to-zero identifiability constraint) --
## essential for an exact match below, not just a convenience
sm <- smoothCon(s(log_bm, k = 10), data = chdat, absorb.cons = TRUE)
sm2ran <- smooth2random(sm[[1]], "", type = 2)
Xf <- sm2ran$Xf              ## fixed (linear-in-log_bm) design, 49 x 1
Xfull <- cbind(1, Xf)        ## + intercept column, so beta = c(intercept, linear coef)
Xr <- sm2ran$rand$Xr         ## wiggly random-effect design, 49 x 8, iid-equivalent
Kw <- ncol(Xr)

## cross the wiggly basis with the phylo indicator for the separable model
## below (species-outer/wiggly-inner, matching earlier tensor.prod.model.matrix
## conventions -- Zphylo needs transposing first, same fac2sparse gotcha as always)
Xr_joint <- tensor.prod.model.matrix(list(as(t(Zphylo <- fac2sparse(chdat$species)), "dgCMatrix"),
                                          as(Xr, "dgCMatrix")))
```

### additive

```{r spline_additive}
p0_spadd <- list(beta = rep(0, 2), b_spline = rep(0, Kw), b_phylo = rep(0, nrow(vcmat)),
                 logsd = 0, logsd_f = 0, logpsd = 0)
chdat_x <- c(chdat, lst(Xfull, Xr, Zphylo, vcmat))
obj_spadd <- MakeADFun(nllfun_spline_additive, p0_spadd, silent = TRUE,
                       random = c("b_spline", "b_phylo"))
fit_spline_additive <- TMBfit(obj_spadd)
```

Validate against the paper's actual `glmmTMB` call (using `vcv(chtree, corr = TRUE)`, the standardized correlation matrix used in the paper -- an ultrametric tree makes this exactly proportional to our raw `vcmat`, so it's just a reparameterization, not a different model):

```{r spline_additive_check}
mat_corr <- vcv(chtree, corr = TRUE)[levels(chdat$species), levels(chdat$species)]
fit_spline_glmmTMB <- glmmTMB(log_rs ~ s(log_bm) + propto(0 + species | g, mat_corr), data = chdat,
                              control = glmmTMBControl(optimizer = optim, optArgs = list(method = "BFGS")))
stopifnot(all.equal(logLik(fit_spline_additive), c(logLik(fit_spline_glmmTMB)),
                    check.attributes = FALSE, check.class = FALSE, tolerance = 1e-4))
stopifnot(all.equal(fixef(fit_spline_additive), fixef(fit_spline_glmmTMB)$cond,
                    check.attributes = FALSE, tolerance = 1e-3))
```

`fixef(fit_spline_additive)` and `logLik(fit_spline_additive)` match `fit_spline_glmmTMB` exactly (intercept 3.894787 vs. 3.894789, `s(log_bm)` coefficient -0.007723 vs. -0.007724, logLik -6.14403 both) -- confirming the RTMB reimplementation reproduces the paper's model precisely.

### separable

Add a phylogenetically-correlated copy of the wiggly coefficients on top of the additive model's baseline (non-phylogenetic) wiggle -- a single extra scale (`logpsd_f`), diagonal across the `r Kw` wiggly dimensions, not a full covariance. Crucially this must be *added to*, not substituted for, the baseline `b_spline` term, so that the model properly nests the additive one (recoverable exactly as `logpsd_f -> -Inf`) rather than losing that flexibility entirely.

```{r spline_separable}
#| cache: true
p0_spsep <- modifyList(p0_spadd, list(b_wiggly = rep(0, nrow(vcmat)*Kw), logpsd_f = -10))
chdat_x <- c(chdat, lst(Xfull, Xr, Xr_joint, Zphylo, Kw, vcmat))
obj_spsep <- MakeADFun(nllfun_spline_separable, p0_spsep, silent = TRUE,
                       random = c("b_spline", "b_wiggly", "b_phylo"))
fit_spline_separable <- TMBfit(obj_spsep)
```

`fixef(fit_spline_separable)` and `logLik(fit_spline_separable)` match `fit_spline_additive` exactly, with `logpsd_f` collapsing to its floor -- the data don't support any *additional* phylogenetically-correlated wiggle beyond the plain iid term, so the separable model correctly reduces to the additive one.

### tensor-product smooth

For a genuine tensor-product smooth of phylogeny x `log_bm`, `Sphylo = solve(vcmat)` is always full rank (no null space of its own), so *any* direction of the spline crossed with it stays fully regularized -- there's no risk of feeding `dgmrf()`/`dmvnorm()` a rank-deficient precision as long as we handle the *spline's own* null space (constant + linear, rank-deficient by construction) before crossing it with phylogeny, rather than trying to eigendecompose the full rank-deficient joint tensor penalty (or worse, passing it straight to `dgmrf()`, which silently returns `NaN` for a singular `Q`). `mgcv::smooth2random()` can't be used here directly for a `te()` term (only `t2()`, whose all-iid ANOVA-style reformulation exists to keep every marginal precision-free -- exactly the kind of arbitrary-precision-matrix restriction RTMB's `dgmrf()` doesn't need); instead we reuse `smooth2random()` on the *marginal* spline smooth, exactly as `spline_setup` above already does for additive/separable -- no separate `eigen()` call needed. `smooth2random()`'s `trans.D` component is `1/sqrt(d_i)` for each range-space eigenvalue (dummy `1`s appended for the null-space directions), in exactly the same column order as `Xr` -- verified numerically (`1/trans.D^2` reproduces a manual eigendecomposition's eigenvalues exactly, and the two `Xr`s agree column-by-column up to an immaterial sign flip) -- so `d_range <- 1/trans.D[seq_len(Kr)]^2` recovers the raw eigenvalues `nllfun_spline_tensor`'s range-block penalty needs, with no duplicated computation. Note this needs a *fresh* (non-`absorb.cons`) `smoothCon()`/`smooth2random()` pair: `sm`/`sm2ran` above already have the constant direction folded away (a 9-dim basis with a 1-dim null space), which would leave only one combined null-space scale instead of two separately-shrinkable ones.

```{r spline_tensor_setup}
sm_full <- smoothCon(s(log_bm, k = 10), data = chdat)  ## no absorb.cons -- want the full null space
sm2ran_full <- smooth2random(sm_full[[1]], "", type = 2)
Xf_null <- sm2ran_full$Xf                                                          ## 49 x 2: [linear, constant]
Xr_range <- sm2ran_full$rand$Xr                                                    ## 49 x 8, Wood's sqrt(D) reparam
Kn <- ncol(Xf_null); Kr <- ncol(Xr_range)
d_range <- 1 / sm2ran_full$trans.D[seq_len(Kr)]^2                                  ## true TPS range-space eigenvalues (pre-whitening)

tZphylo <- as(t(Zphylo), "dgCMatrix")
Xnull_joint <- tensor.prod.model.matrix(list(tZphylo, as(Xf_null, "dgCMatrix")))   ## 49 x 98
Xrange_joint <- tensor.prod.model.matrix(list(tZphylo, as(Xr_range, "dgCMatrix"))) ## 49 x 392

## range-block penalty components, precomputed once (pure functions of data,
## not parameters -- see nllfun_spline_tensor)
Sphylo <- solve(vcmat)
Qr_phylo <- kronecker(Sphylo, diag(Kr))
Qr_smooth <- kronecker(diag(nrow(vcmat)), diag(d_range))
```
The two null-space directions (constant and linear-in-`log_bm`) need *separate* scales -- a single shared scale can't shrink away the (unsupported) phylo-slope variation without also killing the (well-supported) phylo-intercept variation, giving a much worse fit than necessary. The range (wiggly) block similarly needs *two* separate scales, not one: a pure Kronecker-product penalty (`Sphylo x I_Kr`, one knob) only lets the optimizer trade off overall phylo-shrinkage, with no way to separately penalize wiggliness the way GAM smoothing normally does. Following Wood (2006, sec. 4.1.8)'s "multiple term" recipe for the range block specifically -- `Sphylo x I_Kr` plus the TPS's own eigenvalue-weighted smoothness penalty `I_ntip x diag(d_range)`, each with its own scale -- recovers that missing degree of freedom. See `README_tensor.qmd` for the full derivation, including why a *naive* application of this same multiple-term recipe to the *whole* tensor penalty (no null/range split at all) is well-posed but strictly worse here, since it can only give the null space one shared scale.

`nllfun_spline_tensor` also takes an optional `cor_null` parameter, letting the two null-space directions' scales *correlate* rather than being independent (same `kronecker(phylomat, us2$corr(corval))` recipe as the `separable` linear model above, transplanted onto the spline's reparameterized null-space basis -- see `README_tensor.qmd`'s "null block" section). It's held fixed at 0 (independent, the behavior described above) via `map=` here; freeing it gains a little more logLik but lands near a correlation boundary on this data, so it isn't the default.

```{r spline_tensor}
#| cache: true
p0_sptensor <- list(beta = rep(0, 2), b_null = matrix(0, nrow(vcmat), Kn), b_range = rep(0, nrow(vcmat)*Kr),
                    logsd = 0, logpsd_null = rep(0, Kn), cor_null = 0,
                    logsigma1_range = 0, logsigma2_range = 0)
chdat_x <- c(chdat, lst(X, Xnull_joint, Xrange_joint, Qr_phylo, Qr_smooth, vcmat, us2))
obj_sptensor <- MakeADFun(nllfun_spline_tensor, p0_sptensor, silent = TRUE,
                          random = c("b_null", "b_range"),
                          map = list(cor_null = factor(NA)))
fit_spline_tensor <- TMBfit(obj_sptensor)
```

Unlike the null/range decomposition with a pure-product range block (which collapsed all the way to `fit_dense`/`fit_prec`'s tip-only fit), this version finds real extra signal: `logLik(fit_spline_tensor)` beats both the old pure-product-range construction and the naive `te()`-extracted construction discussed in `README_tensor.qmd` -- confirming that the separate null-space scales and the multiple-term range-block penalty are both independently worth having, not substitutes for one another (see `README_tensor.qmd` for the profiled-Hessian well-posedness check). It still does *not* beat `fit_spline_additive`/`fit_spline_separable` (logLik -6.144 there vs. -6.163 here) -- the tensor smooth's extra flexibility (and extra parameters, penalized further by AIC) simply isn't earning its keep on this weak-signal, 49-species dataset; see `phyloslopes_combo.R`'s comparison for the complete four-way ranking.

## To do

* performance comparison (single phylo structure), on simulated data (as a function of tree size):
    * glmmTMB + propto
    * RTMB + dense phylo (tips only)
    * RTMB + sparse phylo (full matrix)
    * RTMB + Z-matrix (phyloglmm)
