---
title: "Tensor-product phylogenetic smooths: naive vs. null/range-decomposed constructions"
author: Ben Bolker / Claude
date: today
format:
  html:
    embed-resources: true
    code-fold: true
---

## Motivation

`phyloslopes.qmd`'s "phylogenetic splines" section fits a genuine tensor-product
smooth of phylogeny x `log_bm` (`nllfun_spline_tensor`, `phyloslopes_utils.R`) by
eigendecomposing the spline's own (rank-deficient) penalty into a null space
(constant + linear) and a range space (wiggly), then giving *each* its own
separate Kronecker-product term crossed with the (always full-rank) phylogenetic
precision. (That describes the construction as it stood when this investigation
started; the range-space block has since changed as a direct result of the
findings below -- see "Recommendation".) `TODO.md` item 8 asked whether this
eigendecomposition is actually necessary, given that RTMB can (in principle)
accept arbitrary precision matrices in `dgmrf()`/`dmvnorm()` -- could we instead
build the tensor-product smooth via `mgcv::smooth.construct()` directly (as in
Clark 2024's blog post) and use its raw `$X`/`$S` output with no decomposition
at all?

This note works through the answer, which turned out to require correcting
several of my own intermediate hypotheses along the way -- kept here so the
false starts (and why they were wrong) are documented, not just the final
answer.

```{r setup, message = FALSE}
library(ade4)
library(ape)
library(dplyr)
library(RTMB)
library(reformulas)
library(mgcv)
library(Matrix)
library(tibble)
library(glmmTMB)
library(numDeriv)
source("phyloslopes_utils.R")

data("carniherbi49", package = "ade4")
chtree <- read.tree(text = carniherbi49$tre2)
chtree$tip.label <- fixup_species(chtree$tip.label)
vcmat <- vcv(chtree)
chdat <- carniherbi49$tab2 |>
  mutate(log_bm = log(bodymass), log_rs = log(runningspeed), g = 1) |>
  tibble::rownames_to_column("species") |>
  mutate(across(species, fixup_species)) |>
  mutate(across(species, \(x) factor(x, levels = rownames(vcmat))))

X <- model.matrix(~ log_bm, data = chdat)
Zphylo <- fac2sparse(chdat$species)
Sphylo <- solve(vcmat)

to_Q <- function(P) {
  if (inherits(P, "advector")) as(as(P, "sparseMatrix"), "adsparse")
  else Matrix::Matrix(P, sparse = TRUE)
}
```

## Two ways to build a tensor-product penalty (Wood 2006, sec. 4.1.8)

Wood's *Generalized Additive Models* distinguishes two ways to combine marginal
penalties $\mathbf S_x, \mathbf S_z$ into a tensor-product smooth's joint penalty
over the full basis:

**"Multiple term penalties"** (what `mgcv::te()` actually uses by default):
$$J(f) \approx \lambda_x J_x^*(f) + \lambda_z J_z^*(f), \qquad
\tilde{\mathbf S}_x = \mathbf S_x \otimes \mathbf I_z, \quad
\tilde{\mathbf S}_z = \mathbf I_x \otimes \mathbf S_z$$
i.e. a **sum** of two Kronecker products, each padded out with an identity in
the *other* dimension.

**"Single penalty"**:
$$\boldsymbol\beta^\top \mathbf S \boldsymbol\beta \quad\text{where}\quad
\mathbf S = \mathbf S_x \otimes \mathbf S_z$$
a literal Kronecker **product** of the two marginal penalties, no identity
padding, no sum.

Wood's verdict on the second option (p. 165), verbatim:

> "but this often leads to **severe undersmoothing**. The reason for the
> undersmoothing is straightforward: **the rank of S is the product of the
> ranks of the S_j**, and in practice this is often far too low for practical
> work. For example, consider a smooth of 3 predictors constructed as a
> tensor product smooth of 3 cubic spline bases, each of rank 5. The
> resulting smooth would have 125 free parameters, but a penalty matrix of
> **rank 27**... By contrast, for the same marginal bases, the **multiple
> term penalties would have rank 117**, leading to a much more useful range
> of effective degrees of freedom."

The mechanism: if $\mathbf S_z$ (say) has a $d$-dimensional null space (e.g.
constant + linear functions, unpenalized by a wiggliness measure), then
$\mathbf S_x \otimes \mathbf S_z$ has null space $\mathbb R^{n_x} \otimes
\text{null}(\mathbf S_z)$ -- an entire $n_x \times d$ block that is
**completely unpenalized** (flat/improper prior), no matter how the
smoothing parameter is tuned. The "multiple term" sum avoids this because
each summand is padded with an identity in the *other* dimension, so as long
as at least one marginal penalty is full rank, the sum is full rank too.

## Why doesn't RTMB choke on a rank-deficient marginal penalty?

It doesn't, because a rank-deficient marginal penalty is never, on its own,
handed to `dgmrf()`. A quick, self-contained illustration:

```{r singular-check}
S2_toy <- matrix(c(2, 0, 0, 0), 2, 2)  ## rank-deficient (1-dim null space)
f_singular <- function(params) -dgmrf(params$b, 0, Q = Matrix::Matrix(S2_toy, sparse = TRUE), log = TRUE)
obj_singular <- MakeADFun(f_singular, list(b = c(0.1, 0.1)), silent = TRUE)
obj_singular$fn(obj_singular$par)  ## NaN: dgmrf() computes its OWN log-determinant for S2_toy alone
```

`S1` (phylo direction) is strictly positive-definite; `S2` (spline direction)
is positive-*semi*-definite (a valid penalty matrix has no negative
eigenvalues). For any finite $\sigma_1, \sigma_2 > 0$,
$$\mathbf x^\top\!\left(\tfrac{1}{\sigma_1^2}\mathbf S_1 + \tfrac{1}{\sigma_2^2}\mathbf S_2\right)\!\mathbf x
= \tfrac{1}{\sigma_1^2}\mathbf x^\top \mathbf S_1 \mathbf x + \tfrac{1}{\sigma_2^2}\mathbf x^\top \mathbf S_2 \mathbf x > 0$$
for every nonzero $\mathbf x$, since the first term alone is already strictly
positive. The **sum** is guaranteed strictly positive-definite regardless of
how deficient $\mathbf S_2$ is by itself -- `dgmrf()` never sees a singular
matrix, only an increasingly ill-conditioned (but always strictly PD) one as
$\sigma_2 \to 0$. Cholesky factorization degrades gracefully under bad
conditioning; it only fails outright at an *exact* zero pivot. So "RTMB
handles a rank-deficient penalty" is really "RTMB never sees a rank-deficient
matrix here at all" -- the multiple-term construction sidesteps the problem
rather than being robust to it.

## Are our `separable` models the (useless) single-penalty case?

No -- both of our `separable` constructions have the algebraic *form* of a
single penalty ($\mathbf Q = \mathbf S_{\text{phylo}} \otimes \mathbf S_{\text{trait}}$,
a literal Kronecker product), but neither triggers Wood's undersmoothing
warning, for two different reasons:

- **`nllfun_sep` / `nllfun_dense_slopes`** (linear intercept+slope): the
  trait factor is a full $2\times 2$ covariance matrix from `unstructured(2)`,
  parameterized (log-sds + a correlation transform) to be positive-definite
  for *any* parameter value. It's a genuine covariance over a 2-dimensional
  trait space, not a wiggliness *penalty* -- nothing about (intercept, slope)
  is "smoother" than anything else, so there's no null space to begin with.
  $\text{rank}(\mathbf S_{\text{phylo}} \otimes \mathbf S_{\text{trait}}^{-1}) = 49 \times 2 = 98$,
  full rank on the nose.
- **`nllfun_spline_separable`**: the phylo-wiggle term multiplies `Sphylo` by
  `Xr`, the *whitened* range-space design from `smooth2random()`'s
  $1/\sqrt{d_i}$ reparameterization -- a scaled identity, full rank by
  construction. The null space wasn't absent here (a spline's wiggliness
  penalty genuinely has one); it was stripped out beforehand and handled
  separately, via the plain iid `b_spline` baseline term.

The actual useless single-penalty case would be $\mathbf S_{\text{phylo}}
\otimes \mathbf S_{\text{spline,raw}}$ -- the *raw*, un-whitened, rank-8-of-10
TPS penalty, multiplied directly with no preprocessing. We've never built
that as a standalone model; it only ever appears as one summand inside the
naive multiple-term sum below, where `S1` alone keeps the total full rank.
Built alone it would be both singular (the `NaN` demonstrated above) and, if
patched to avoid that, would have a genuinely free 98-dimensional null space
-- Wood's exact undersmoothing scenario.

## Building the naive (multiple-term) tensor smooth directly from `smooth.construct()`

Per the original TODO question: yes, we can extract `$X`/`$S` directly from a
`te()` smooth, no eigendecomposition, and reuse the *already-generic*
`nllfun_tensor` (`phyloslopes_utils.R`, originally written for the trivial
`diag(2)` random-slopes case) unchanged:

```{r naive-tensor}
te_spec <- te(species, log_bm, bs = c("mrf", "tp"), k = c(49, 10),
             xt = list(list(penalty = Sphylo), NULL))
sm_te <- smooth.construct(te_spec, data = chdat, knots = NULL)
tX_naive <- sm_te$X
S1 <- sm_te$S[[1]]; S2 <- sm_te$S[[2]]
cat("dim(tX_naive):", dim(tX_naive), "\n")
cat("rank(S1):", Matrix::rankMatrix(S1), " rank(S2):", Matrix::rankMatrix(S2),
   "(out of", nrow(S1), ")\n")

chdat_x <- c(chdat, lst(X, tX = tX_naive, tp1 = Matrix(S1, sparse = TRUE), tp2 = Matrix(S2, sparse = TRUE)))
p0_naive <- list(beta = rep(0, 2), b = rep(0, ncol(tX_naive)), logsigma1 = 0, logsigma2 = 0, logsd_resid = 0)
obj_naive <- MakeADFun(nllfun_tensor, p0_naive, silent = TRUE, random = "b")
fit_naive <- nlminb(obj_naive$par, obj_naive$fn, obj_naive$gr,
                    control = list(rel.tol = 1e-13, eval.max = 3000, iter.max = 3000))
H_naive <- numDeriv::jacobian(obj_naive$gr, fit_naive$par)
ev_naive <- eigen(H_naive, symmetric = TRUE, only.values = TRUE)$values

cat("convergence:", fit_naive$convergence, fit_naive$message, "\n")
cat("logLik:", -fit_naive$objective, "\n")
cat("Hessian eigenvalues:", round(ev_naive, 4), "\n")
```

Clean convergence, no `NaN`, essentially no negative curvature (the smallest
eigenvalue is at the edge of numerical noise, not a genuine saddle like the
`diag(2)` pathology documented in `phyloslopes_tiny_explore.R`). This
confirms the tiny-example finding (naive Kronecker-sum is well-identified
for a *genuine* TPS penalty, unlike the trivial `diag(2)` trait case)
generalizes to the real 49-species data.

But it gives a **different fit** from the (then-current) null/range-decomposed
`nllfun_spline_tensor` construction, not the same one reparameterized -- for
comparison, that construction as it stood at the start of this investigation
(separate null-space scales, but a single-scale *pure* Kronecker-product range
block). `nllfun_spline_tensor` itself has since been changed to the "hybrid"
range block derived below, so that old, single-scale range-block version is
reproduced here as a standalone local function (`nllfun_old_pure_product`)
purely so it can still appear as a comparison row below -- it is deliberately
*not* calling `nllfun_spline_tensor`, which now implements something else:

```{r current-tensor}
## smooth2random(), not a separate eigendecomposition, for the null/range
## split (see "Building the naive..." above and phyloslopes_utils.R) --
## its trans.D is 1/sqrt(eigenvalue) per range-space direction, in the same
## column order as Xr
sm_full <- smoothCon(s(log_bm, k = 10), data = chdat)  ## no absorb.cons -- want the full null space
sm2ran_full <- smooth2random(sm_full[[1]], "", type = 2)
Xf_null <- sm2ran_full$Xf
Xr_range <- sm2ran_full$rand$Xr
Kn <- ncol(Xf_null); Kr <- ncol(Xr_range)
d_range <- 1 / sm2ran_full$trans.D[seq_len(Kr)]^2  ## true TPS range-space eigenvalues

tZphylo <- as(t(Zphylo), "dgCMatrix")
Xnull_joint <- tensor.prod.model.matrix(list(tZphylo, as(Xf_null, "dgCMatrix")))
Xrange_joint <- tensor.prod.model.matrix(list(tZphylo, as(Xr_range, "dgCMatrix")))

nllfun_old_pure_product <- function(params) {
  getAll(params, chdat_x)
  mu <- drop(X %*% beta + Xnull_joint %*% b_null + Xrange_joint %*% b_range)
  REPORT(mu); ADREPORT(mu)
  resid <- log_rs - mu; REPORT(resid)
  Sigma_unsc_null <- kronecker(vcmat, diag(length(logpsd_null)))
  pen_null <- -dmvnorm(b_null, rep(0, length(b_null)), Sigma = Sigma_unsc_null,
                        scale = rep(exp(logpsd_null), nrow(vcmat)), log = TRUE)
  Q_range <- (1/exp(logpsd_range)^2) * kronecker(Sphylo, diag(Kr))
  pen_range <- -dgmrf(b_range, 0, Q = to_Q(Q_range), log = TRUE)
  lik <- -sum(dnorm(log_rs, mean = mu, sd = exp(logsd), log = TRUE))
  lik + pen_null + pen_range
}
chdat_x <- c(chdat, lst(X, Xnull_joint, Xrange_joint, Kr, vcmat))
p0_old <- list(beta = rep(0, 2), b_null = rep(0, nrow(vcmat)*Kn), b_range = rep(0, nrow(vcmat)*Kr),
              logsd = 0, logpsd_null = rep(0, Kn), logpsd_range = 0)
obj_old <- MakeADFun(nllfun_old_pure_product, p0_old, silent = TRUE, random = c("b_null", "b_range"))
fit_current <- nlminb(obj_old$par, obj_old$fn, obj_old$gr,
                      control = list(rel.tol = 1e-13, eval.max = 3000, iter.max = 3000))
cat("old pure-product-range logLik:", -fit_current$objective, "\n")
cat("naive (te()-extracted) logLik:", -fit_naive$objective, "\n")
```

## Validating `te()` + `nllfun_tensor` against an independent worked example

Before trusting `nllfun_tensor`'s behavior on our own (as it turns out,
severely overparameterized -- see below) 49-species data, it's worth
checking the basic machinery agrees with `mgcv::gam()` itself on a case
`gam()` can actually fit. @clarkPhylogeneticSmoothing2024's own worked
example does exactly this: a `te()` smooth crossing an 8-level `mrf` time
margin with a 12-species `mrf` phylogenetic margin (`k = c(8, 12)`, 96
total tensor-product coefficients), on 450 complete-case observations (12
species x 50 time points, minus two withheld species and the last 5 time
points per species) -- small and well-conditioned, unlike our own case.

`clark_2024_ex.R` reproduces that example verbatim (down to fitting
`gam()` and timing it: well under a second). Reusing its simulated `dat`,
`omega` (phylogenetic precision), `rw_penalty`, and `N_species` here:

```{r clark-setup, message = FALSE}
source("clark_2024_ex.R")
```

For a direct, apples-to-apples comparison of the `te()`-handling itself,
this drops Clark's additional `s(time, k = 10)` term (fitting the `te()`
component alone against `y`) and works with complete cases only, so both
`gam()` and RTMB see identical rows/basis:

```{r clark-compare}
dat_cc <- dat[stats::complete.cases(dat[, c("y", "time_factor", "species")]), ]
te_spec <- te(time_factor, species, bs = c("mrf", "mrf"), k = c(8, N_species),
             xt = list(list(penalty = rw_penalty), list(penalty = omega)))
sm_te <- smooth.construct(te_spec, data = dat_cc, knots = NULL)
tX <- sm_te$X; S1 <- sm_te$S[[1]]; S2 <- sm_te$S[[2]]

## RTMB (nllfun_tensor, unmodified). X_clark, not X: X is already the
## 49-species model.matrix(~log_bm, chdat) from earlier sections, and is
## reused (from the global environment) by chdat_x in "Isolating the
## mechanism" below -- overwriting it here would silently break that
X_clark <- matrix(1, nrow(dat_cc), 1)  ## intercept-only fixed effect
chdat_x <- lst(log_rs = dat_cc$y, X = X_clark, tX, tp1 = Matrix(S1, sparse = TRUE), tp2 = Matrix(S2, sparse = TRUE))
p0 <- list(beta = 0, b = rep(0, ncol(tX)), logsigma1 = 0, logsigma2 = 0, logsd_resid = 0)
obj_rtmb <- MakeADFun(nllfun_tensor, p0, silent = TRUE, random = "b")
t_rtmb <- system.time({
  fit_rtmb <- nlminb(obj_rtmb$par, obj_rtmb$fn, obj_rtmb$gr,
                     control = list(rel.tol = 1e-12, eval.max = 2000, iter.max = 2000))
})
pl_rtmb <- obj_rtmb$env$parList(par = obj_rtmb$env$last.par.best)
mu_rtmb <- drop(X_clark %*% fit_rtmb$par["beta"] + tX %*% pl_rtmb$b)

## gam(), same te() spec, same complete-case data
mod_ml <- gam(y ~ te(time_factor, species, bs = c("mrf", "mrf"), k = c(8, N_species),
                     xt = list(list(penalty = rw_penalty), list(penalty = omega))),
             data = dat_cc, drop.unused.levels = FALSE, method = "ML")
mod_reml <- update(mod_ml, method = "REML")

data.frame(
  quantity = c("elapsed time (s)", "logLik", "intercept"),
  RTMB = c(t_rtmb["elapsed"], -fit_rtmb$objective, unname(fit_rtmb$par["beta"])),
  gam_ML = c(NA, logLik(mod_ml), unname(coef(mod_ml)[1])),
  gam_REML = c(NA, logLik(mod_reml), unname(coef(mod_reml)[1]))
)
cat("max|mu_rtmb - fitted(mod_ml)|:  ", max(abs(mu_rtmb - fitted(mod_ml))), "\n")
cat("cor(mu_rtmb, fitted(mod_ml)):   ", cor(mu_rtmb, fitted(mod_ml)), "\n")
cat("recovered sigma_resid:", exp(fit_rtmb$par["logsd_resid"]),
   " (true simulated sd = 0.35)\n")
```

Fitted values agree essentially exactly (correlation > 0.99999, max
difference ~0.002 -- consistent with the two optimizers, `nlminb` vs.
`gam()`'s own PIRLS+Newton, converging to slightly different points on a
very flat final region rather than a real disagreement), and the recovered
residual SD lands right on the true simulated value (0.35). This confirms
`te()` + `nllfun_tensor` and `mgcv::gam()`'s own tensor-product handling
are fitting the theoretically/conceptually equivalent model.

The raw `logLik` values do **not** match (RTMB's Laplace-integrated
marginal likelihood and `mgcv`'s REML/ML-reported `logLik()` use different
normalizing-constant conventions -- a known caveat when comparing
TMB-based and mgcv-based likelihoods directly, not a sign of disagreement)
-- what's reparameterization-invariant (fitted values, recovered variance
components) is exactly what matches.

This also explains, by contrast, why the same check isn't attempted
directly on our own 49-species data: `gam()` doesn't converge on it at
all. Diagnosis (not reproduced here): the `mrf(49) x tp(10)` tensor basis
(490 coefficients on 49 observations, one per species, no replication) is
so overparameterized that the fit interpolates the data almost immediately
(deviance reaches ~1e-19, i.e. numerically zero, within the first couple
of outer iterations even at a smaller `k = c(49, 4)`) -- once deviance is
numerically zero, `mgcv`'s Newton-based smoothing-parameter selection has
no well-defined gradient to converge on (a whole flat manifold of
smoothing parameters gives the identical perfect fit) and never
terminates. This is the same single-observation-per-tip overparameterization
pathology documented at length in `phyloslopes_tiny_explore.R`, just
surfacing as *non-convergence* in `mgcv`'s Newton iteration rather than as
*boundary collapse* in RTMB's variance-component estimates -- our
Laplace-approximation route tolerates this specific degeneracy far more
gracefully than `mgcv`'s REML/ML machinery does.

## Isolating the mechanism

Two structurally *independent* questions are tangled up inside "naive vs.
current": how the null (constant+linear) block is scaled, and how the range
(wiggle) block is penalized. Untangling them:

```{r mechanism, message = FALSE}
## (1) null-space block ALONE, one shared scale for both directions
nllfun_nullspace_shared <- function(params) {
  getAll(params, chdat_x)
  mu <- drop(X %*% beta + tX %*% b)
  REPORT(mu); ADREPORT(mu)
  resid <- log_rs - mu; REPORT(resid)
  Sigma_full <- kronecker(vcmat, diag(rep(exp(2*logpsd_shared), Kn), nrow = Kn))
  pen <- -dmvnorm(b, rep(0, length(b)), Sigma = Sigma_full, log = TRUE)
  lik <- -sum(dnorm(log_rs, mean = mu, sd = exp(logsd_resid), log = TRUE))
  lik + pen
}
chdat_x <- c(chdat, lst(X, tX = Xnull_joint, vcmat, Kn))
p0 <- list(beta = rep(0, 2), b = rep(0, ncol(Xnull_joint)), logpsd_shared = 0, logsd_resid = 0)
fit_null_shared <- nlminb(MakeADFun(nllfun_nullspace_shared, p0, silent = TRUE, random = "b")$par,
                          MakeADFun(nllfun_nullspace_shared, p0, silent = TRUE, random = "b")$fn,
                          MakeADFun(nllfun_nullspace_shared, p0, silent = TRUE, random = "b")$gr,
                          control = list(rel.tol = 1e-13, eval.max = 3000, iter.max = 3000))

## (2) null-space block ALONE, separate scales (nllfun_tensor_nullspace)
p0b <- list(beta = rep(0, 2), b = rep(0, ncol(Xnull_joint)), logpsd_null = rep(0, Kn), logsd_resid = 0)
obj_null_sep <- MakeADFun(nllfun_tensor_nullspace, p0b, silent = TRUE, random = "b")
fit_null_sep <- nlminb(obj_null_sep$par, obj_null_sep$fn, obj_null_sep$gr,
                       control = list(rel.tol = 1e-13, eval.max = 3000, iter.max = 3000))

## (3) hybrid: SAME separate null scales as the old construction above, but
## the range block rebuilt the multiple-term way (Sphylo x I + I x
## diag(true TPS eigenvalues)) instead of the old pure Kronecker product
## (Sphylo x I alone). This is now exactly what nllfun_spline_tensor itself
## implements (see Recommendation below) -- called directly here, not a
## local copy, so this comparison also doubles as a check that the real
## function reproduces the derivation
Qr_phylo <- kronecker(Sphylo, diag(Kr))
Qr_smooth <- kronecker(diag(nrow(vcmat)), diag(d_range))
chdat_x <- c(chdat, lst(X, Xnull_joint, Xrange_joint, Qr_phylo, Qr_smooth, vcmat))
p0c <- list(beta = rep(0, 2), b_null = rep(0, nrow(vcmat)*Kn), b_range = rep(0, nrow(vcmat)*Kr),
           logsd = 0, logpsd_null = rep(0, Kn), logsigma1_range = 0, logsigma2_range = 0)
obj_hybrid <- MakeADFun(nllfun_spline_tensor, p0c, silent = TRUE, random = c("b_null", "b_range"))
fit_hybrid <- nlminb(obj_hybrid$par, obj_hybrid$fn, obj_hybrid$gr,
                     control = list(rel.tol = 1e-13, eval.max = 3000, iter.max = 3000))

data.frame(
  model = c("null-only, 1 shared scale", "null-only, separate scales",
           "old (separate null + pure-product range)",
           "hybrid == current nllfun_spline_tensor (separate null + multiple-term range)",
           "naive te() (1 shared null + multiple-term range)"),
  logLik = round(c(-fit_null_shared$objective, -fit_null_sep$objective,
                   -fit_current$objective, -fit_hybrid$objective, -fit_naive$objective), 4)
)
```

This cleanly separates two independent effects:

1. **The null block wants separate per-direction scales.** Giving intercept
   and slope their own phylogenetic scale (rather than sharing one) is worth
   ~4.4 nats on its own (null-only: shared vs. separate). The multiple-term
   recipe *cannot* provide this: `S2`'s null block is exactly zero there, so
   only `sigma1` ever touches the null space, forcing one shared scale on
   both directions no matter how the rest of the model is built.
2. **The range block wants the multiple-term structure**, independently of
   the null-block question. Holding the (already-separated) null block fixed
   and rebuilding *only* the range block as `Sphylo %x% I + I %x%
   diag(d_range)` instead of `Sphylo %x% I` alone gains another ~0.49 nats
   (old vs. hybrid). This is Wood's point, correctly scoped: the range
   block's own smoothness structure (the TPS's real eigenvalues) gives the
   optimizer a second, phylo-independent knob for trading off wiggliness
   against fit -- a pure Kronecker product can only "shrink everything
   toward phylogenetic similarity or don't," with no way to separately
   control smoothness.

The naive `te()` fit sits between the two extremes: it gets the multiple-term
range benefit but loses the separate-null-scales benefit (one shared scale,
forced). The **hybrid** construction -- keep `nllfun_spline_tensor`'s
separate null-space scales, but rebuild its range block as a multiple-term
sum instead of a pure product -- beats every other *tensor-smooth*
construction tried, including the fully-naive one.

It does **not**, however, beat the additive/separable models
(`phyloslopes_combo.R`): `fit_spline_additive`/`fit_spline_separable` both
get logLik $-6.144$, still better than the hybrid tensor's $-6.163$ (and by
AIC, more so, since the tensor smooth has more parameters to be penalized
for). On this weak-signal, 49-species dataset, letting the linear/wiggle
trend vary phylogenetically at all (tensor's whole premise) just isn't
earning its keep relative to the simpler additive structure -- the fixes
here are about making the tensor-smooth *family* well-posed and as
strong as it can be, not about making it win the model comparison.

## Recommendation

`nllfun_spline_tensor` is not "the same model as the naive `te()` one, just
implemented more carefully to avoid `NaN`" -- it's a genuinely different
(and, on the null-space side, better-identified) construction. But its
range-space block was leaving real fit on the table by using a pure
Kronecker product where a multiple-term sum does strictly better, per
Wood's own argument about effective-degrees-of-freedom range. Implemented:
`nllfun_spline_tensor`'s range-space penalty is now `(1/sigma1^2)
* kron(Sphylo, I_Kr) + (1/sigma2^2) * kron(I_ntip, diag(d_range))` (the
`hybrid` construction above), keeping the separate null-space scales
unchanged, rather than switching wholesale to the naive `te()`-extracted
version (which would sacrifice the separate null-space scales that are worth
more than the range-space fix on their own).
