---
title: "analyzing vignettes with a mixed ordinal model"
date: today
author: "Ben Bolker"
bibliography: clmm_sing.bib
format:
  html:
    code-fold: true
    code-summary: "Show the code"
    embed-resources: true
---

## Options

```{r opts}
focal_question <- 1
fn  <- "clmm_fits_Q1.rds"
ncores <- 12
```

```{r pkgs, message = FALSE}
library(readxl)
library(ordinal)
library(broom.mixed)
library(parallel)
library(ggplot2); theme_set(theme_bw())
library(dplyr)  ## mutate/filter/etc.
library(purrr)
library(emmeans)
## library(dotwhisker) ## not used [raw ggplot instead]
## library(marginaleffects) ## doesn't work for clmm() models yet ...
```

Read and transform data:

```{r read-data}
## dataset <- read_excel("C:/Users/Anmol/Documents/suicide stats 25/r SUICIDE/Suicide Study 24 25 FINAL.xlsx")
infn <- "SUICIDE STAT CLEANED FOR THESIS.xlsx"
if (file.exists(infn)) {
  dataset <- read_excel(infn)
}
#making sure the file is not empty
if (!is.null(dataset)) {
  dataset <- transform(dataset,
                       RESP = factor(RESP, 
                                     levels = 1:4,
                                     labels = c("Strongly Disagree", "Disagree", "Agree", "Strongly Agree"),
                                     ordered = TRUE),
  TB = factor(TB),
  PB = factor(PB),
  LV = factor(LV),
  ## construct interactions by hand, clmm() can't handle these
  PB_LV_TB = interaction(PB, LV, TB), #main eff
  RESPID_PB_LV_TB = interaction(RESPID, PB, LV, TB),
  DE_PB_LV_TB = interaction(DE, PB, LV, TB)
  )
}
```

Fit all of the models (or recover the saved fits from a file). 

- the fully crossed maximal model [@barrRandom2013], including variation of three-way interaction and all lower-order terms across both `RESPID` and `DE`
- drop three-way interactions (only main effects and two-way interactions)
- main effect varying only
- `CRI`: "complex random interactions" *sensu* @scandolaReliabilityFeasibilityLinear2024
- `CRI_minus`: intercept-only for `DE`, CRI for `RESPID`; this is based on using `isSingular(., collapse=FALSE)` (as defined below) to see which particular terms of the model above were singular

```{r fit-models}
if (file.exists(fn)) {
  fits <- readRDS(fn)
} else {
  
  ## we only have 4 jobs to run (so far), but each one is multi-threaded, so
  ##   need more cores in cluster ... ???
  ##
  ## i have 12 codes
  cl <- makeCluster(min(ncores, detectCores()-1))

  ## all possible (so far) choices for random effect terms
  re_forms <- c(
    ## fully crossed (maximal model)
    full = "(PB*LV*TB|RESPID) + (PB*LV*TB|DE)",
    ## two-way interactions among factors
    ##  (only reduces from 8x8 to 7x7 covariance matrix)
    twoway = "((PB+LV+TB)^2|RESPID) + ((PB+LV+TB)^2|DE)", #incl this cuz if we r looking at 3 way, we must incl all the lower order ones
    ## main effects only
    maineff = "(PB+LV+TB|RESPID) + (PB+LV+TB|DE)",
    ## complex random intercepts à la Scandola and Tidoni
    ## (need to spell these out, clmm can't expand in-place)
    CRI = "(1|RESPID) + (1|RESPID_PB_LV_TB) + (1|DE) + (1|DE_PB_LV_TB)",
    CRI_minus = "(1|RESPID) + (1|DE) + (1|DE_PB_LV_TB)"
  )

  fitfun <- function(re_form, question = 1) { #change question based on the q analyzing (1-3)
    ## add fixed effects (full three-way interaction) to specified RE
    form <- reformulate(c("PB*LV*TB", re_form), response = "RESP") #reformulate creates a formula from a character vector
    dd <- subset(dataset, QUES == question)
    t1 <- system.time(
      fit <- clmm(form, #form = formula
                  data = dd,
                  ## 'trace' gives noisy output (but disappears if run in parallel)
                  control = clmm.control(trace = 1,
                                         iter.max = 1000, eval.max = 1000)
                  )
    )
    attr(fit, "time") <- t1
    fit
  }

  ## fit everything in parallel - have to load packages and export data
  invisible(clusterEvalQ(cl, library(ordinal)))
  clusterExport(cl, c("dataset", "re_forms"))
  fits <- parLapply(cl, re_forms, fitfun)

  stopCluster(cl)
  saveRDS(fits, file = fn)
}
```

```{r}
if (exists("dd")) {
  dataQ <- subset(dd, question == focal_question)
  ## checking distribution/experimental design
  with(dataQ, table(table(PB, LV, TB, RESPID)))  ## 2 observations for every combination (because we had one that was parent vs partner)
  ## for example ...
  subset(dataQ, LV ==1 & TB == 0 & PB == 1 & RESPID ==84)
}
```

Both RE terms from the full model are singular (determinant approx 0)

```{r}
sapply(VarCorr(fits[["full"]]), det)
```

Check if any of the RE terms is singular (zero variance etc.: see [GLMM FAQ](https://bbolker.github.io/mixedmodels-misc/glmmFAQ.html#singular-fits), [lme4 documentation](https://lme4.r-universe.dev/lme4/doc/manual.html#isSingular))

```{r sing-vals-funs}
isSingular <- function(x, tol = 1e-8, collapse = TRUE) {
  dets <- vapply(VarCorr(x), det, FUN.VALUE = numeric(1))
  sing <- dets < tol
  if (!collapse) sing else any(sing)
}
get_time <- function(x, w = "elapsed") {
  attr(x, "time")[[w]]
}

AICtab <- purrr::map_dfr(fits,
                         function(x) {
                           tibble(AIC = AIC(x),
                                  sing = isSingular(x))
                         },
                         .id = "model") |>
  mutate(delta_AIC = AIC-min(AIC), .after = "model") |>
  select(-AIC) |>
  arrange(delta_AIC)
```

```{r print-aictab}
knitr::kable(AICtab, digits=1)
```

Choosing RE (random effects) on the basis of minimum AIC is reasonable ... e.g. @batesParsimonious2015, @matuschekBalancing2017, @moritzrole2023a

Timings (in seconds):
```{r times}
sapply(fits, get_time, "sys.self")  
```

Proceed with simplest (lowest-AIC, non-singular) model:

```{r}
fCRI <- fits[["CRI_minus"]]
```

```{r eval=FALSE, include = FALSE}
get_sds <- function(fit) {
  sapply(VarCorr(fit), attr, "stddev")
}
print(get_sds(fCRI))
```

<!-- I'm not sure why last two variances are identical - that may be because of balanced experimental design or ??
-->

Would normally worry about scaling by 2SD to draw coefficient plots [@gelmanScaling2008; @schielzethSimple2010], but all predictors are categorical/binary, so should be comparable:

```{r fit-funs}
## first two fits give non-pos-def Hessians, have to work harder
## (move this fix upstream to the `tidy` method?)
tfun <- function(x, alpha = 0.05) {
  npd <- inherits(try(vcov(x), silent = TRUE), "try-error")
  if (npd) {
    ## do tidying by hand, using `Matrix::nearPD` to force positive definite
    ##  Hessian/allow SE computation
    vv <- Matrix::nearPD(x$Hessian)$mat
    se <- sqrt(diag(solve(vv)))
    cc <- coef(x)
    ## | indicates intercepts, ST are variances
    omitfun <- function(x) {
      x[!grepl("(ST|\\|)", names(x))]
    }
    cc <- omitfun(cc)
    se <- omitfun(se)
    qq <- qnorm(1-alpha/2)
    ret <- tibble(term = names(cc),
                  estimate = cc,
                  std.error = se,
                  statistic = cc/se,
                  p.value = 2*pnorm(-abs(statistic)),
                  lwr = cc - qq*se, upr = cc + qq*se,
                  )
  } else {
    ret <- tidy(x, effects = "fixed", conf.int = TRUE)  |>
      filter(coef.type == "location") |>
      rename(lwr = "conf.low", upr = "conf.high")
  }
  ret
}
```

```{r sum-fits}
reorder_fun <- function(x) {
  x <- mutate(x, across(term, ~ reorder(factor(.), estimate)))
  if ("model" %in% names(x)) {
    x <- mutate(x, across(model, forcats::fct_inorder))
  }
  x
}
tt_all <- purrr::map_dfr(fits, tfun, .id = "model") |> reorder_fun()
tt <- tfun(fCRI) |> reorder_fun()
```

```{r plot-gg-all, message=FALSE}
gg_all <- ggplot(tt_all, aes(estimate, term, color = model)) +
  geom_pointrange(aes(xmin=lwr, xmax = upr),
                  position = position_dodge(width = 0.3)) +
  geom_vline(xintercept = 0, lty = 2)

print(gg_all)
ggsave(gg_all, file = "param_comparison.pdf")
```

These are in units of log-odds.

Final model only (could use `dotwhisker::dwplot()` but I like to tweak things a little bit more)

```{r final-model-coefplot}
gg_final <- ggplot(tt, aes(estimate, term)) +
  geom_pointrange(aes(xmin=lwr, xmax = upr)) +
  geom_vline(xintercept = 0, lty = 2)
print(gg_final)
```

Note for interpretation purposes that these effects are using *treatment contrasts* (see below)

Can look at expected marginal means, for every level:

```{r emmeans}
plot(emmeans(fCRI, ~PB*LV*TB))
```

or for one particular main effect

```{r emmeans-PB}
(em_PB <- emmeans(fCRI, ~PB))
pairs(em_PB)
```

Refit with sum-to-zero contrasts. This doesn't change the model, but it changes the interpretation of the lower-level terms (main effects and two-way interactions). These are now the *average* effects across levels rather than effect at the baseline level. For example, in the previous (default, treatment contrasts) model, the `PB` effect was the effect of `PB` at the baseline levels of the other variables (`TB=0`, `DE=0`). In the model below, it is the average of the effect across the combinations of other variables. Normally we could use `emmeans` to do this without fitting the model, but because we are fitting a specialized random effects models,  we need to refit.

```{r fit-zero-sum}
clist <- rep("contr.sum", 3) |> setNames(c("TB", "LV", "PB")) |> as.list()
## have to do some fancy stuff because the model was fitted within a function,
## the objects we need to re-fit the model are no longer visible ...
form <- formula(fCRI)
dd <- subset(dataset, QUES == focal_question)
fCRI2 <- update(fCRI,
                contrasts = clist,
                control = clmm.control()  ## turn off tracing etc.
                )
tt2 <- tfun(fCRI2) |>  reorder_fun()
suppressWarnings(gg_zerocontrasts <- gg_final %+% tt2)
```

We get a fairly different set of parameters (this means the *interpretation*
is different under these contrasts -- the overall model itself is the same)
have to decide which interpretation is actually the most meaningful for this
example ...

Coefficient plot:

```{r zero-contrast-plot}
print(gg_zerocontrasts)
```

```{r coefmat}
tt2 |>
  rename("p-value" = "p.value") |>
  select(-c(term, lwr, upr)) |>
  as.data.frame() |>  ## convert to data frame before setting rownames
  `rownames<-`(tt2$term) |>
  printCoefmat(digits = 2)
```

Estimates and confidence intervals:

```{r}
tt2 |> select(term, estimate, lwr, upr) |> knitr::kable(digits=3)
```
