---
author: Ben Bolker
title: "How should we model combinations of phylogenetic correlations with other (non-scalar) random effects models?"
date: today
format:
    html:
       embed-resources: true
       code-fold: true
---

Tow (potential) parts to this:

* benchmarking alternative approaches to phylogenetic fits (Z-matrix, dense (tip-only) precision matrix, sparse (full tree) precision matrix ...
* alternative approaches to crossed (phylogenetic × something else): additive, separable, tensor-product

```{r pkgs, message=FALSE}
if (!require("MRFtools")) {
    stop('please remotes::install_github("eric-pedersen/MRFtools")')
    ##remotes::install_github("eric-pedersen/MRFtools")
}
if (!require("phyloglmm")) {
    stop('please remotes::install_github("wzmli/phyloglmm/pkg")')
}

## data / data manipulation
library(ade4)
library(ape)
library(dplyr)
## modeling
library(glmmTMB)
library(RTMB)
library(reformulas)
library(mgcv)
library(MRFtools)
library(Matrix)
## model diagnostics, etc.
library(performance)
library(ggplot2); theme_set(theme_bw())
library(cowplot)
library(broom.mixed) ## for tidy()
```

The McGillycuddy *et al.* manuscript on 'propto' random effects (i.e. random effects where the covariance matrix is known *a priori* up to a scalar multiplier) illustrates how to use `glmmTMB` + `propto` to fit models with a phylogenetic term.

```{r utils}
source("phyloslopes_utils.R")
```

```{r setup}
data("carniherbi49", package = "ade4")
chtree <- read.tree(text=carniherbi49$tre2)
chtree$tip.label <- fixup_species(chtree$tip.label)
vcmat <- vcv(chtree)
chdat <- carniherbi49$tab2 |>
  mutate(log_bm = log(bodymass),
         log_rs = log(runningspeed),
         g = 1  ## dummy for grouping variable
         ) |>
  tibble::rownames_to_column("species") |>
  mutate(across(species, fixup_species)) |>
  mutate(across(species, \(x) factor(x, levels = rownames(vcmat))))
## check
stopifnot(identical(sort(rownames(vcmat)), sort(as.character(chdat$species))))
save(chdat, chtree, vcmat, file = "phyloslopes.rda")
```

Model fit from McGillycuddy et al.:
```{r fit1}
fit_propto_glmmTMB <- glmmTMB(log_rs ~ log_bm + propto(0 + species | g, vcmat),
              data = chdat)
```

```{r fit1_check, eval=FALSE}
performance::check_model(fit_propto_glmmTMB,
                         check = c("linearity",
                                   "homogeneity", "pp_check",
                                   "qq"))
```

RTMB analogue:

```{r rtmb1}
X <- model.matrix(~ log_bm, data = chdat)
Z <- fac2sparse(chdat$species)
## for intercept-slope model
us2 <- unstructured(2)
rt <- mkReTrms(findbars(~ (1 + log_bm | species)),
               fr = chdat, calc.lambdat = FALSE)
## move data into required wrapper (no extras needed for this version)
chdat_x <- chdat
p0 <- list(beta = rep(0, 2), b = rep(0, nrow(vcmat)), logsd=0, logpsd = 0)
obj <- MakeADFun(nllfun1, p0, silent = TRUE, random = "b")
fit_propto_TMB <- TMBfit(obj)
```

```{r glmmTMB_rtmb_check}
stopifnot(all.equal(fixef(fit_propto_TMB), fixef(fit_propto_glmmTMB)$cond,
                    check.attributes = FALSE, tolerance = 1e-7))
```

```{r phylo_prec}
## Matrix(solve(vcmat)) auto-detects symmetry and returns a *dense*
## "dsyMatrix" -- not a "sparseMatrix" subclass, so it can't be coerced
## to "adsparse" (RTMB's AD sparse-matrix type), nor to "dgCMatrix"
## (note also the capital C -- "dgcMatrix" isn't a real class name).
## sparse = TRUE forces a sparse ("dsCMatrix") representation instead,
## which *is* a "sparseMatrix" and coerces fine.
chdat_x <- c(chdat, list(phyloprec =
                           Matrix(solve(vcmat), sparse = TRUE)))
obj <- MakeADFun(nllfun_prec, p0, silent = TRUE, random = "b")
```

Precision matrix over the *whole* tree (tips + internal nodes), via `MRFtools::mrf_penalty(..., internal_nodes = TRUE)`. The full `ntip + Nnode` precision matrix is rank-deficient by 1: the classic Brownian-motion "no fixed origin" singularity, since shifting every node (including the root) by the same constant leaves all edge contrasts unchanged. Dropping the root's row/column from the precision matrix (and the corresponding column from `Z`) removes the singularity:

```{r phylo_prec_allnodes}
ntip <- length(chtree$tip.label)
Q_full_raw <- mrf_penalty(chtree, "brownian", internal_nodes = TRUE)
## mrf_penalty() orders [tips][internal nodes]; reorder to [internal][tips]
## to match the factor levels used to build Z_full below, so that the root
## (the first internal node) ends up at position 1
ord <- c(ntip + seq_len(chtree$Nnode), seq_len(ntip))
Q_full <- as(Q_full_raw[ord, ord], "CsparseMatrix")
Q_noroot <- Q_full[-1, -1]

## fac2sparse() returns levels x observations (the "Zt" convention); use
## placeholder levels 1:Nnode for the (never-observed -> all-zero-column)
## internal nodes, then transpose to the usual observations x levels form
Z_full <- t(fac2sparse(factor(chdat$species,
                              levels = c(seq_len(chtree$Nnode), chtree$tip.label)),
                       drop.unused.levels = FALSE))
Z_noroot <- Z_full[, -1]

p0_noroot <- modifyList(p0, list(b = rep(0, chtree$Nnode + ntip - 1)))
chdat_x <- c(chdat, list(Z = Z_noroot, phyloprec = Q_noroot))
obj_noroot <- MakeADFun(nllfun_prec, p0_noroot, silent = TRUE, random = "b")
fit_prec_allnodes_noroot <- TMBfit(obj_noroot)
```

`fixef(fit_prec_allnodes_noroot)` and `logLik(fit_prec_allnodes_noroot)` come out identical to `fit_dense`/`fit_prec` -- i.e. marginalizing the internal nodes really does reduce to the tip-only model, confirming the `phyloglmm`-equivalence conjecture from the intro, once the root is pinned down.

A third way to encode the same tree structure: `phylo.to.Z()` builds a tip x edge matrix with a `sqrt(edge.length)` entry wherever an edge is ancestral to a tip, so `Z %*% b` with `b` iid unit-scale reproduces `vcv(chtree)` exactly -- no covariance/precision matrix needed, and (unlike the all-nodes precision matrix above) no root-singularity to work around, since there's no free "root value" parameter.

```{r phylo_edge}
pZ <- phylo.to.Z(chtree)
Z_edge <- pZ[as.character(chdat$species), ]

p0_edge <- modifyList(p0, list(b = rep(0, nrow(chtree$edge))))
chdat_x <- c(chdat, list(Z = Z_edge))
obj_edge <- MakeADFun(nllfun_edge, p0_edge, silent = TRUE, random = "b")
fit_edge <- TMBfit(obj_edge)
```

`fixef(fit_edge)` and `logLik(fit_edge)` again match `fit_dense`/`fit_prec`/`fit_prec_allnodes_noroot`, confirming the `phyloglmm` `Z`-matrix equivalence a second, more direct way.

Thought I would try adding `clade` as a fixed effect, but it doesn't seem to
do much --- abandon that avenue for now.

## separable

`Z = t(rt$Zt)` orders its columns `[species-outer, trait-inner]` (each species' intercept/slope pair adjacent), so `b` (an `ntip x 2` matrix) must be flattened to match via `c(t(b))`, not `c(b)` -- flattening column-major would silently permute `b` relative to `Z`'s columns (trait-outer/species-inner vs. species-outer/trait-inner) and land the optimizer on a much worse fit. `nllfun_sep` in `phyloslopes_utils.R` has this fix.

```{r fit_sep1}
p2 <- list(beta = rep(0, 2), b = matrix(0,
                                        nrow = nrow(vcmat),
                                        ncol = 2),
           logrsd=rep(0, 2),
           logsdres = 0,
           corval=0)
chdat_x <- c(chdat, list(phylomat = vcmat, scale = 1, Z = t(rt$Zt)))
obj2 <- MakeADFun(nllfun_sep, p2, silent = TRUE, random = "b")
fit_sep <- TMBfit(obj2)
```

As a ground-truth check, we can also write out the *dense* Kronecker covariance directly: stack the 49 species x 2 traits into one 98-vector `b` with `b ~ N(0, Sigma2 %x% vcmat)`, no `dseparable()` abstraction involved.

```{r dense_slopes}
Zdense <- t(rt$Zt)
p0_ds <- modifyList(p0, list(b = rep(0, 2*nrow(vcmat)), logrsd = rep(0,2), corval = 0))
chdat_x <- c(chdat, list(Zdense = Zdense))
obj_ds <- MakeADFun(nllfun_dense_slopes, p0_ds, silent = TRUE, random = "b")
fit_dense_slopes <- TMBfit(obj_ds)
```

`fixef(fit_sep)` and `fixef(fit_dense_slopes)` now agree (as do their `logLik`s), confirming `dseparable()` really does implement the `Sigma2 %x% vcmat` covariance.

Now the edge-based version, answering the to-do question below ("does the Z-matrix random-slopes model match the separable model?"): `KhatriRao(t(pZ), t(J))` builds a `nobs x (2*nedge)` matrix whose columns are ordered `[edge-outer, trait-inner]` -- each edge contributes an (intercept-innovation, slope-innovation) pair. Propagating an edge-level, iid-across-edges, 2x2-correlated Brownian innovation up the tree induces `Cov(trait_a_i, trait_b_j) = Sigma2[a,b] * vcmat[i,j]` for species `i,j` -- exactly the same separable covariance as above, just built without ever forming `vcmat` or a covariance/precision matrix at all.

```{r phylo_edge_slopes}
f <- ~ 1 + (1 + log_bm | species)
J <- eval(bquote(model.matrix(~.(findbars(f)[[1]][[2]]), data = chdat)))
pZ <- phylo.to.Z(chtree)
## reorder pZ's tip-label rows to match chdat's (species-factor) row order
pZ_ord <- pZ[as.character(chdat$species), ]
KR <- t(KhatriRao(t(pZ_ord), t(J)))

p0_es <- modifyList(p0, list(b = rep(0, 2*nrow(chtree$edge)), logrsd = rep(0,2), corval = 0))
chdat_x <- c(chdat, list(KR = KR))
obj_es <- MakeADFun(nllfun_edge_slopes, p0_es, silent = TRUE, random = "b")
fit_edge_slopes <- TMBfit(obj_es)
```

`fixef(fit_edge_slopes)` and `logLik(fit_edge_slopes)` match `fit_sep`/`fit_dense_slopes` too (all three agree to several significant figures) -- so yes, the edge-based `Z`-matrix random-slopes model is exactly the separable model, just reparameterized.

## tensor-product

```{r tp_tests}
tp <- tensor.prod.penalties(list(diag(2), solve(vcmat)))
lapply(tp, dim)
plot_grid(plotlist = lapply(tp, ifun))
tX <- tensor.prod.model.matrix(list(Z, as(X, "dgCMatrix")))
dim(tX)
```

I believe based on Wood (2006) that we need the penalty to be the sum of the scales/penalty values times (`penalty %*% b`), with
the `b` reused in each case. (Does the `scale` parameter of `RTMB::dgmrf` apply in the opposite direction, i.e. do we need
`RTMB::dgmrf(S, scale = 1/sd)` ?

I think in this case it's OK that the null space is empty - we do want this 'random effect' to disappear completely
as the penalty goes to infinity.

**Is the separable random-slopes model (above) nested in this tensor-product model?** No. For `b` laid out species-outer/trait-inner (so the separable covariance is `Sigma_full = vcmat %x% Sigma2`), the separable model's precision is a Kronecker *product*
$$Q_{sep} = S_{phylo} \otimes P_{trait,sep},$$
while summing `-dgmrf()` over `tensor.prod.penalties(list(Sphylo, Ptrait))` gives a Kronecker *sum*
$$Q_{add} = S_{phylo} \otimes I_2 \;+\; I_{ntip} \otimes P_{trait,add}.$$
Both are diagonalized by the same rotation $V \otimes I_2$, where $S_{phylo} = V\Lambda V^\top$ ($\Lambda=\mathrm{diag}(\lambda_1,\dots,\lambda_{49})$). In that basis, block $i$ of $Q_{sep}$ is $\lambda_i P_{trait,sep}$ and block $i$ of $Q_{add}$ is $\lambda_i I_2 + P_{trait,add}$. Matching every block simultaneously requires
$$P_{trait,add} = \lambda_i\,(P_{trait,sep} - I_2) \quad \text{for all } i = 1,\dots,49,$$
which is impossible unless $P_{trait,sep} = I_2$ (which forces $P_{trait,add}=0$, not a valid precision matrix), since $S_{phylo}$'s eigenvalues are not all equal (range $\approx 0.0032$-$3.33$ for this tree, i.e. $S_{phylo}$ is not proportional to $I_{ntip}$). So the separable model is *not* a special case of this tensor-product-penalty model -- Kronecker product and Kronecker sum are different families of precision matrices, and no choice of the free 2x2 trait-precision parameters lets the tensor-product model exactly reproduce the separable one.

This matches what happens numerically: fitting this model (`Sphylo` fixed/unscaled, `Ptrait` built from 3 free parameters -- two log-sds and a correlation, then inverted) converges to a much worse fit (logLik $\approx -180$ vs. $-6.5$ for the separable model) and lands on a genuine saddle point -- the profiled Hessian (via `numDeriv::jacobian(obj$gr, ...)`) has substantial negative eigenvalues (e.g. $-3011,-107$ from an all-zero start; $-55,-7.3$ from a warm start at the separable model's optimum) -- regardless of how the trait correlation is parameterized (`unstructured(2)`'s internal transform vs. the glmmTMB-vignette mapping $\rho=\theta/\sqrt{1+\theta^2}$ give near-identical results). The optimizer keeps pushing the correlation toward its boundary, trying (and failing) to approximate the unreachable separable structure.

Clark, Nicholas. “Phylogenetic Smoothing Using Mgcv.” GAMbler, 2024. https://ecogambler.netlify.app/blog/phylogenetic-smooths-mgcv/.

```{r mrf_penalty_test}
m1 <- mrf_penalty(chtree, "brownian", internal_nodes = FALSE)
m2 <- mrf_penalty(chtree, "brownian", internal_nodes = TRUE)
image(Matrix(zapsmall(m1)))
dim(zapsmall(m1))
dim(m2)
```

If using all-nodes parameterization of phylogenetic precision matrix (to take advantage of sparsity/`gmrf()` in `RTMB`), we'd need to make sure that the 'species' factor included values (not included in data set) for the internal nodes ...

## phylogenetic splines

McGillycuddy *et al.*'s `propto` manuscript (`main.tex`) fits an *additive* model combining a penalized thin-plate-spline smoother of `log_bm` with a phylogenetic random intercept:
```
glmmTMB(log(runningspeed) ~ s(logbm) + propto(0 + species | g, mat), data = data,
        control = glmmTMBControl(optimizer = optim, optArgs = list(method = "BFGS")))
```
the two random effects (spline wiggle, phylogenetic intercept) are completely independent, summed in the linear predictor. Here we redo this comparing three structures: (1) that additive model, (2) a *separable* model where the spline's wiggly coefficients also get their own (single-scale) phylogenetically-correlated component, and (3) a genuine tensor-product smooth of phylogeny x `log_bm`.

```{r spline_setup}
## thin-plate spline marginal for log_bm; absorb.cons=TRUE matches glmmTMB's
## internal s() handling (its own sum-to-zero identifiability constraint) --
## essential for an exact match below, not just a convenience
sm <- smoothCon(s(log_bm, k = 10), data = chdat, absorb.cons = TRUE)
sm2ran <- smooth2random(sm[[1]], "", type = 2)
Xf <- sm2ran$Xf              ## fixed (linear-in-log_bm) design, 49 x 1
Xfull <- cbind(1, Xf)        ## + intercept column, so beta = c(intercept, linear coef)
Xr <- sm2ran$rand$Xr         ## wiggly random-effect design, 49 x 8, iid-equivalent
Kw <- ncol(Xr)

## cross the wiggly basis with the phylo indicator for the separable model
## below (species-outer/wiggly-inner, matching earlier tensor.prod.model.matrix
## conventions -- Zphylo needs transposing first, same fac2sparse gotcha as always)
Xr_joint <- tensor.prod.model.matrix(list(as(t(Zphylo <- fac2sparse(chdat$species)), "dgCMatrix"),
                                          as(Xr, "dgCMatrix")))
```

### additive

```{r spline_additive}
p0_spadd <- list(beta = rep(0, 2), b_spline = rep(0, Kw), b_phylo = rep(0, nrow(vcmat)),
                 logsd = 0, logsd_f = 0, logpsd = 0)
chdat_x <- chdat
obj_spadd <- MakeADFun(nllfun_spline_additive, p0_spadd, silent = TRUE,
                       random = c("b_spline", "b_phylo"))
fit_spline_additive <- TMBfit(obj_spadd)
```

Validate against the paper's actual `glmmTMB` call (using `vcv(chtree, corr = TRUE)`, the standardized correlation matrix used in the paper -- an ultrametric tree makes this exactly proportional to our raw `vcmat`, so it's just a reparameterization, not a different model):

```{r spline_additive_check}
mat_corr <- vcv(chtree, corr = TRUE)[levels(chdat$species), levels(chdat$species)]
fit_spline_glmmTMB <- glmmTMB(log_rs ~ s(log_bm) + propto(0 + species | g, mat_corr), data = chdat,
                              control = glmmTMBControl(optimizer = optim, optArgs = list(method = "BFGS")))
stopifnot(all.equal(logLik(fit_spline_additive), c(logLik(fit_spline_glmmTMB)),
                    check.attributes = FALSE, tolerance = 1e-4))
stopifnot(all.equal(fixef(fit_spline_additive), fixef(fit_spline_glmmTMB)$cond,
                    check.attributes = FALSE, tolerance = 1e-3))
```

`fixef(fit_spline_additive)` and `logLik(fit_spline_additive)` match `fit_spline_glmmTMB` exactly (intercept 3.894787 vs. 3.894789, `s(log_bm)` coefficient -0.007723 vs. -0.007724, logLik -6.14403 both) -- confirming the RTMB reimplementation reproduces the paper's model precisely.

### separable

Add a phylogenetically-correlated copy of the wiggly coefficients on top of the additive model's baseline (non-phylogenetic) wiggle -- a single extra scale (`logpsd_f`), diagonal across the `r Kw` wiggly dimensions, not a full covariance. Crucially this must be *added to*, not substituted for, the baseline `b_spline` term, so that the model properly nests the additive one (recoverable exactly as `logpsd_f -> -Inf`) rather than losing that flexibility entirely.

```{r spline_separable}
p0_spsep <- modifyList(p0_spadd, list(b_wiggly = rep(0, nrow(vcmat)*Kw), logpsd_f = -10))
chdat_x <- c(chdat, list(Xr_joint = Xr_joint))
obj_spsep <- MakeADFun(nllfun_spline_separable, p0_spsep, silent = TRUE,
                       random = c("b_spline", "b_wiggly", "b_phylo"))
fit_spline_separable <- TMBfit(obj_spsep)
```

`fixef(fit_spline_separable)` and `logLik(fit_spline_separable)` match `fit_spline_additive` exactly, with `logpsd_f` collapsing to its floor -- the data don't support any *additional* phylogenetically-correlated wiggle beyond the plain iid term, so the separable model correctly reduces to the additive one.

### tensor-product smooth

For a genuine tensor-product smooth of phylogeny x `log_bm`, `Sphylo = solve(vcmat)` is always full rank (no null space of its own), so *any* direction of the spline crossed with it stays fully regularized -- there's no risk of feeding `dgmrf()`/`dmvnorm()` a rank-deficient precision as long as we handle the *spline's own* null space (constant + linear, rank-deficient by construction) before crossing it with phylogeny, rather than trying to eigendecompose the full rank-deficient joint tensor penalty (or worse, passing it straight to `dgmrf()`, which silently returns `NaN` for a singular `Q`). `mgcv::smooth2random()` can't be used here directly for a `te()` term (only `t2()`, whose all-iid ANOVA-style reformulation exists to keep every marginal precision-free -- exactly the kind of arbitrary-precision-matrix restriction RTMB's `dgmrf()` doesn't need); instead we eigendecompose the spline's own 10x10 penalty directly. Note this needs a *fresh* (non-`absorb.cons`) `smoothCon()` call: `sm` above already has the constant direction folded away (a 9-dim basis with a 1-dim null space), which would leave only one combined null-space scale instead of two separately-shrinkable ones.

```{r spline_tensor_setup}
sm_full <- smoothCon(s(log_bm, k = 10), data = chdat)  ## no absorb.cons -- want the full null space
S_spline <- sm_full[[1]]$S[[1]]
ev <- eigen(S_spline, symmetric = TRUE)
null_idx <- which(ev$values < 1e-8 * max(ev$values))
range_idx <- which(ev$values >= 1e-8 * max(ev$values))
Xf_null <- sm_full[[1]]$X %*% ev$vectors[, null_idx, drop = FALSE]                 ## 49 x 2: [linear, constant]
Xr_range <- sm_full[[1]]$X %*% ev$vectors[, range_idx, drop = FALSE] %*%
  diag(1/sqrt(ev$values[range_idx]))                                              ## 49 x 8, Wood's sqrt(D) reparam
Kn <- ncol(Xf_null); Kr <- ncol(Xr_range)

tZphylo <- as(t(Zphylo), "dgCMatrix")
Xnull_joint <- tensor.prod.model.matrix(list(tZphylo, as(Xf_null, "dgCMatrix")))   ## 49 x 98
Xrange_joint <- tensor.prod.model.matrix(list(tZphylo, as(Xr_range, "dgCMatrix"))) ## 49 x 392
```

The two null-space directions (constant and linear-in-`log_bm`) need *separate* scales -- a single shared scale can't shrink away the (unsupported) phylo-slope variation without also killing the (well-supported) phylo-intercept variation, giving a much worse fit than necessary:

```{r spline_tensor}
p0_sptensor <- list(beta = rep(0, 2), b_null = rep(0, nrow(vcmat)*Kn), b_range = rep(0, nrow(vcmat)*Kr),
                    logsd = 0, logpsd_null = rep(0, Kn), logpsd_range = 0)
chdat_x <- c(chdat, list(Xnull_joint = Xnull_joint, Xrange_joint = Xrange_joint))
obj_sptensor <- MakeADFun(nllfun_spline_tensor, p0_sptensor, silent = TRUE,
                          random = c("b_null", "b_range"))
fit_spline_tensor <- TMBfit(obj_sptensor)
```

`fixef(fit_spline_tensor)` and `logLik(fit_spline_tensor)` match `fit_dense`/`fit_prec` (the very first tip-only phylo-intercept models) exactly -- both extra components (phylo-slope, phylo-wiggle) shrink cleanly to zero for this weak-signal dataset, and (unlike the earlier random-slopes tensor-product exploration) the profiled Hessian has no negative eigenvalues, just the boundary-flat directions expected from variance components at zero -- a well-posed fit, not a saddle.

## To do

* ~~test cov (`dmvnorm`) implementation vs precision (`dgmrf`) implementation in RTMB~~ done, see `phylo_prec`/`glmmTMB_rtmb_check`
* ~~extend to phylo + smooths (need to separate fixed-effect params) [additive, separable, tensor-product]~~ done, see "phylogenetic splines" section above
* performance comparison (single phylo structure), on simulated data (as a function of tree size):
    * glmmTMB + propto
    * RTMB + dense phylo (tips only)
    * RTMB + sparse phylo (full matrix)
    * RTMB + Z-matrix (phyloglmm)
* which model does the Z-matrix with random slopes correspond to? (What is the implied correlation between species-A intercept and species-B slope? Does it match the separable model?) i.e. what is $\mathbf Z \boldsymbol \Sigma  \mathbf Z^\top$ (analytically or numerically) ... ?
* try out tensor-product model

```{r pic1}
ggplot(chdat, aes(runningspeed, bodymass, colour = clade)) +
  geom_point(size=4) +
  scale_x_log10() +
  scale_y_log10() +
  scale_colour_manual(values = c("darkred", "darkblue"))
```
