---
author: Ben Bolker
title: "How should we model combinations of phylogenetic correlations with other (non-scalar) random effects models?"
date: today
format:
    html:
       embed-resources: true
       code-fold: true
---

Tow (potential) parts to this:

* benchmarking alternative approaches to phylogenetic fits (Z-matrix, dense (tip-only) precision matrix, sparse (full tree) precision matrix ...
* alternative approaches to crossed (phylogenetic × something else): additive, separable, tensor-product

```{r pkgs, message=FALSE}
if (!require("MRFtools")) {
    stop('please remotes::install_github("eric-pedersen/MRFtools")')
    ##remotes::install_github("eric-pedersen/MRFtools")
}
if (!require("phyloglmm")) {
    stop('please remotes::install_github("wzmli/phyloglmm/pkg")')
}

## data / data manipulation
library(ade4)
library(ape)
library(dplyr)
## modeling
library(glmmTMB)
library(RTMB)
library(reformulas)
library(mgcv)
library(MRFtools)
library(Matrix)
## model diagnostics, etc.
library(performance)
library(ggplot2); theme_set(theme_bw())
library(cowplot)
library(broom.mixed) ## for tidy()
```

The McGillycuddy *et al.* manuscript on 'propto' random effects (i.e. random effects where the covariance matrix is known *a priori* up to a scalar multiplier) illustrates how to use `glmmTMB` + `propto` to fit models with a phylogenetic term.

```{r utils}
source("phyloslopes_utils.R")
```

```{r setup}
data("carniherbi49", package = "ade4")
chtree <- read.tree(text=carniherbi49$tre2)
chtree$tip.label <- fixup_species(chtree$tip.label)
vcmat <- vcv(chtree)
chdat <- carniherbi49$tab2 |>
  mutate(log_bm = log(bodymass),
         log_rs = log(runningspeed),
         g = 1  ## dummy for grouping variable
         ) |>
  tibble::rownames_to_column("species") |>
  mutate(across(species, fixup_species)) |>
  mutate(across(species, \(x) factor(x, levels = rownames(vcmat))))
## check
stopifnot(identical(sort(rownames(vcmat)), sort(as.character(chdat$species))))
save(chdat, chtree, vcmat, file = "phyloslopes.rda")
```

Model fit from McGillycuddy et al.:
```{r fit1}
fit_propto_glmmTMB <- glmmTMB(log_rs ~ log_bm + propto(0 + species | g, vcmat),
              data = chdat)
```

```{r fit1_check, eval=FALSE}
performance::check_model(fit_propto_glmmTMB,
                         check = c("linearity",
                                   "homogeneity", "pp_check",
                                   "qq"))
```

RTMB analogue:

```{r rtmb1}
X <- model.matrix(~ log_bm, data = chdat)
Z <- fac2sparse(chdat$species)
## for intercept-slope model
us2 <- unstructured(2)
rt <- mkReTrms(findbars(~ (1 + log_bm | species)),
               fr = chdat, calc.lambdat = FALSE)
## move data into required wrapper (no extras needed for this version)
chdat_x <- chdat
p0 <- list(beta = rep(0, 2), b = rep(0, nrow(vcmat)), logsd=0, logpsd = 0)
obj <- MakeADFun(nllfun1, p0, silent = TRUE, random = "b")
fit_propto_TMB <- TMBfit(obj)
```

```{r glmmTMB_rtmb_check}
stopifnot(all.equal(fixef(fit_propto_TMB), fixef(fit_propto_glmmTMB)$cond,
                    check.attributes = FALSE, tolerance = 1e-7))
```

```{r phylo_prec}
## Matrix(solve(vcmat)) auto-detects symmetry and returns a *dense*
## "dsyMatrix" -- not a "sparseMatrix" subclass, so it can't be coerced
## to "adsparse" (RTMB's AD sparse-matrix type), nor to "dgCMatrix"
## (note also the capital C -- "dgcMatrix" isn't a real class name).
## sparse = TRUE forces a sparse ("dsCMatrix") representation instead,
## which *is* a "sparseMatrix" and coerces fine.
chdat_x <- c(chdat, list(phyloprec =
                           Matrix(solve(vcmat), sparse = TRUE)))
obj <- MakeADFun(nllfun_prec, p0, silent = TRUE, random = "b")
```

Precision matrix over the *whole* tree (tips + internal nodes), via `MRFtools::mrf_penalty(..., internal_nodes = TRUE)`. The full `ntip + Nnode` precision matrix is rank-deficient by 1: the classic Brownian-motion "no fixed origin" singularity, since shifting every node (including the root) by the same constant leaves all edge contrasts unchanged. Dropping the root's row/column from the precision matrix (and the corresponding column from `Z`) removes the singularity:

```{r phylo_prec_allnodes}
ntip <- length(chtree$tip.label)
Q_full_raw <- mrf_penalty(chtree, "brownian", internal_nodes = TRUE)
## mrf_penalty() orders [tips][internal nodes]; reorder to [internal][tips]
## to match the factor levels used to build Z_full below, so that the root
## (the first internal node) ends up at position 1
ord <- c(ntip + seq_len(chtree$Nnode), seq_len(ntip))
Q_full <- as(Q_full_raw[ord, ord], "CsparseMatrix")
Q_noroot <- Q_full[-1, -1]

## fac2sparse() returns levels x observations (the "Zt" convention); use
## placeholder levels 1:Nnode for the (never-observed -> all-zero-column)
## internal nodes, then transpose to the usual observations x levels form
Z_full <- t(fac2sparse(factor(chdat$species,
                              levels = c(seq_len(chtree$Nnode), chtree$tip.label)),
                       drop.unused.levels = FALSE))
Z_noroot <- Z_full[, -1]

p0_noroot <- modifyList(p0, list(b = rep(0, chtree$Nnode + ntip - 1)))
chdat_x <- c(chdat, list(Z = Z_noroot, phyloprec = Q_noroot))
obj_noroot <- MakeADFun(nllfun_prec, p0_noroot, silent = TRUE, random = "b")
fit_prec_allnodes_noroot <- TMBfit(obj_noroot)
```

`fixef(fit_prec_allnodes_noroot)` and `logLik(fit_prec_allnodes_noroot)` come out identical to `fit_dense`/`fit_prec` -- i.e. marginalizing the internal nodes really does reduce to the tip-only model, confirming the `phyloglmm`-equivalence conjecture from the intro, once the root is pinned down.

Thought I would try adding `clade` as a fixed effect, but it doesn't seem to
do much --- abandon that avenue for now.

## separable

```{r fit_sep1}
p2 <- list(beta = rep(0, 2), b = matrix(0,
                                        nrow = nrow(vcmat),
                                        ncol = 2),
           logrsd=rep(0, 2),
           logsdres = 0,
           corval=0)
chdat_x <- c(chdat, list(phylomat = vcmat, scale = 1, Z = t(rt$Zt)))
obj2 <- MakeADFun(nllfun_sep, p2, silent = TRUE, random = "b")
fit_sep <- TMBfit(obj2)
```

```{r test_phyloZ}
f <- ~ 1 + (1+log_bm | species)
J <- eval(bquote(model.matrix(~.(findbars(f)[[1]][[2]]), data = chdat)))
pZ <- phylo.to.Z(chtree)
KR <- t(KhatriRao(t(pZ), t(J)))
image(KR)
```

## tensor-product

```{r tp_tests}
tp <- tensor.prod.penalties(list(diag(2), solve(vcmat)))
lapply(tp, dim)
plot_grid(plotlist = lapply(tp, ifun))
tX <- tensor.prod.model.matrix(list(Z, as(X, "dgCMatrix")))
dim(tX)
```

I believe based on Wood (2006) that we need the penalty to be the sum of the scales/penalty values times (`penalty %*% b`), with
the `b` reused in each case. (Does the `scale` parameter of `RTMB::dgmrf` apply in the opposite direction, i.e. do we need
`RTMB::dgmrf(S, scale = 1/sd)` ?

I think in this case it's OK that the null space is empty - we do want this 'random effect' to disappear completely 
as the penalty goes to infinity.


Clark, Nicholas. “Phylogenetic Smoothing Using Mgcv.” GAMbler, 2024. https://ecogambler.netlify.app/blog/phylogenetic-smooths-mgcv/.

```{r mrf_penalty_test}
m1 <- mrf_penalty(chtree, "brownian", internal_nodes = FALSE)
m2 <- mrf_penalty(chtree, "brownian", internal_nodes = TRUE)
image(Matrix(zapsmall(m1)))
dim(zapsmall(m1))
dim(m2)
```

If using all-nodes parameterization of phylogenetic precision matrix (to take advantage of sparsity/`gmrf()` in `RTMB`), we'd need to make sure that the 'species' factor included values (not included in data set) for the internal nodes ...

## To do

* test cov (`dmvnorm`) implementation vs precision (`dgmrf`) implementation in RTMB
* extend to phylo + smooths (need to separate fixed-effect params) [additive, separable, tensor-product]
* performance comparison (single phylo structure), on simulated data (as a function of tree size):
    * glmmTMB + propto
    * RTMB + dense phylo (tips only)
    * RTMB + sparse phylo (full matrix)
    * RTMB + Z-matrix (phyloglmm)
* which model does the Z-matrix with random slopes correspond to? (What is the implied correlation between species-A intercept and species-B slope? Does it match the separable model?) i.e. what is $\mathbf Z \boldsymbol \Sigma  \mathbf Z^\top$ (analytically or numerically) ... ?
* try out tensor-product model

```{r pic1}
ggplot(chdat, aes(runningspeed, bodymass, colour = clade)) +
  geom_point(size=4) +
  scale_x_log10() +
  scale_y_log10() +
  scale_colour_manual(values = c("darkred", "darkblue"))
```
