---
title: "How should we model combinations of phylogenetic correlations with other (non-scalar) random effects models?"
---

```{r pkgs}
## remotes::install_github("eric-pedersen/MRFtools")
library(MRFtools)
library(ade4)
library(glmmTMB)
library(ape)
library(Matrix)
library(ggplot2); theme_set(theme_bw())
library(dplyr)
```

The McGillycuddy *et al.* manuscript on 'propto' random effects (i.e. random effects where the covariance matrix is known *a priori* up to a scalar multiplier) illustrates how to use `glmmTMB` + `propto` to fit models with a phylogenetic term.

```{r setup}
gsub2 <- function(x, pattern, replacement) {
  gsub(pattern, replacement, x)
}
fixup_species <- function(x) {
  x |>
    gsub2("\\.", "_") |>
    gsub2("nelsoni$", "") |>
    gsub2("cenereorenteus", "cinereoargenteus") |>
    gsub2("hemonius", "hemionus") |>
    gsub2("thomsonii", "thompsonii")
}

data("carniherbi49", package = "ade4")
chdat <- carniherbi49$tab2 |>
  mutate(log_bm = log(bodymass),
         log_rs = log(runningspeed),
         g = 1  ## dummy for grouping variable
         ) |>
  tibble::rownames_to_column("species") |>
  mutate(across(species, fixup_species))
## FIXME
vcmat <- vcv(chtree)

setdiff(chdat$species, chtree$tip.label)
setdiff(chtree$tip.label, chdat$species)
chtree <- read.tree(text=carniherbi49$tre2)
m1 <- glmmTMB(log_rs ~ log_bm + propto(0 + species | g, vcmat),
              data = chdat)
```

e.g.

* phylogenetic × random-slopes
* phylogenetic × smooth terms (penalized regression splines)

## additive

## separable

## tensor-product

how does constructing a tensor product smooth actually work? (We need to combine both bases and penalty matrices, right?)

cf. `phyloglmm` (which does this by putting everything in the $Z$ matrix -- should be equivalent, can we prove it?

```{r}
install_github("eric-pedersen/MRFtools")
```

Clark, Nicholas. “Phylogenetic Smoothing Using Mgcv.” GAMbler, 2024. https://ecogambler.netlify.app/blog/phylogenetic-smooths-mgcv/.


```{r}
remotes::install_github("eric-pedersen/MRFtools")
library(MRFtools)
library(ade4)
library(glmmTMB)
library(ape)
library(Matrix)
library(ggplot2); theme_set(theme_bw())
data("carniherbi49")
chdat <- carniherbi49$tab2
chtree <- read.tree(text=carniherbi49$tre2)

## 
m1 <- mrf_penalty(chtree, "brownian", internal_nodes = FALSE)
m2 <- mrf_penalty(chtree, "brownian", internal_nodes = TRUE)
image(Matrix(zapsmall(m1)))
dim(zapsmall(m1))
dim(m2)
```

If using all-nodes parameterization of phylogenetic precision matrix (to take advantage of sparsity/`gmrf()` in `RTMB`), we'd need to make sure that the 'species' factor included values (not included in data set) for the internal nodes ...

## To do

* replicate examples from `propto` paper, in RTMB (linear and functional effects of log(body mass) on running speed): additive model
* try out separable model via `dseparable`
* try out tensor-product model
* what happens if we include a fixed effect of clade?

```{r pic1}
ggplot(chdat, aes(runningspeed, bodymass, colour = clade)) +
  geom_point(size=4) +
  scale_x_log10() +
  scale_y_log10() +
  scale_colour_manual(values = c("darkred", "darkblue"))
```
