---
title: "How should we model combinations of phylogenetic correlations with other (non-scalar) random effects models?"
---

```{r pkgs}
## remotes::install_github("eric-pedersen/MRFtools")
library(MRFtools)
library(ade4)
library(glmmTMB)
library(ape)
library(Matrix)
library(ggplot2); theme_set(theme_bw())
library(dplyr)
library(RTMB)
library(reformulas)
```

The McGillycuddy *et al.* manuscript on 'propto' random effects (i.e. random effects where the covariance matrix is known *a priori* up to a scalar multiplier) illustrates how to use `glmmTMB` + `propto` to fit models with a phylogenetic term.

```{r setup}
gsub2 <- function(x, pattern, replacement) {
  gsub(pattern, replacement, x)
}

fixup_species <- function(x) {
  x |>
    gsub2("\\.", "_") |>
    gsub2("nelsoni$", "") |>
    gsub2("cinereorenteus", "cinereoargenteus") |>
    gsub2("hemonius", "hemionus") |>
    gsub2("thomsonii", "thompsonii") |>
    gsub2("Odicoileus", "Odocoileus") |>
    gsub2("_$", "")
}

data("carniherbi49", package = "ade4")
chtree <- read.tree(text=carniherbi49$tre2)
chtree$tip.label <- fixup_species(chtree$tip.label)
vcmat <- vcv(chtree)
chdat <- carniherbi49$tab2 |>
  mutate(log_bm = log(bodymass),
         log_rs = log(runningspeed),
         g = 1  ## dummy for grouping variable
         ) |>
  tibble::rownames_to_column("species") |>
  mutate(across(species, fixup_species)) |>
  mutate(across(species, \(x) factor(x, levels = rownames(vcmat))))


stopifnot(identical(sort(rownames(vcmat)), sort(chdat$species)))

m1 <- glmmTMB(log_rs ~ log_bm + propto(0 + species | g, vcmat),
              data = chdat)
```

RTMB analogue:

```{r}
X <- model.matrix(~ log_bm, data = chdat)
Z <- fac2sparse(chdat$species)
nllfun1 <- function(params) {
  getAll(params, chdat)
  mu <- X %*% beta + Z %*% b
  pen <- -dmvnorm(b, 0, Sigma = vcmat, scale = exp(logpsd), log = TRUE)
  lik <- -sum(dnorm(log_rs, mean = mu, sd = exp(logsd), log = TRUE))
  lik + pen
}
p0 <- list(beta = rep(0, 2), b = rep(0, nrow(vcmat)), logsd=0, logpsd = 0)
obj <- MakeADFun(nllfun1, p0, silent = TRUE, random = "b")
fit <- with(obj, nlminb(par, fn, gr))
```

```{r}
stopifnot(all.equal(obj$env$parList()$beta, fixef(m1)$cond,
                    check.attributes = FALSE, tolerance = 1e-7))
```

(clade doesn't interfere with phylog signal??)
```{r}
m2 <- update(m1, . ~ . + clade)
```

## separable

```{r}
us2 <- unstructured(2)
rt <- mkReTrms(findbars(~ (1 + log_bm | species)),
               fr = chdat, calc.lambdat = FALSE) # <2>
Z_sep <- t(rt$Zt)
nllfun_sep <- function(params) {
  getAll(params, chdat)
  mu <- X %*% beta + Z_sep %*% c(b)
  f_phylo <- function(x) dmvnorm(x, 0, Sigma = vcmat, scale = 1, log = TRUE)
  f_cov <- function(x) dmvnorm(x, 0,
                               Sigma = us2$corr(corval), scale = exp(logsd))
  pen <- -dseparable(f_phylo, f_cov)(b)
  lik <- -sum(dnorm(log_rs, mean = mu, sd = exp(logsd), log = TRUE))
  lik + pen
}
p2 <- list(beta = rep(0, 2), b = matrix(0,
                                        nrow = nrow(vcmat),
                                        ncol = 2), logsd=rep(0, 2),
           corval=0)
obj2 <- MakeADFun(nllfun_sep, p2, silent = TRUE, random = "b")
fit_sep <- with(obj2, nlminb(par, fn, gr))
```
Not working yet ...

## tensor-product

how does constructing a tensor product smooth actually work? (We need to combine both bases and penalty matrices, right?)

cf. `phyloglmm` (which does this by putting everything in the $Z$ matrix -- should be equivalent, can we prove it?

```{r}
install_github("eric-pedersen/MRFtools")
```

Clark, Nicholas. “Phylogenetic Smoothing Using Mgcv.” GAMbler, 2024. https://ecogambler.netlify.app/blog/phylogenetic-smooths-mgcv/.


```{r}
remotes::install_github("eric-pedersen/MRFtools")
library(MRFtools)
library(ade4)
library(glmmTMB)
library(ape)
library(Matrix)
library(ggplot2); theme_set(theme_bw())
data("carniherbi49")
chdat <- carniherbi49$tab2
chtree <- read.tree(text=carniherbi49$tre2)

## 
m1 <- mrf_penalty(chtree, "brownian", internal_nodes = FALSE)
m2 <- mrf_penalty(chtree, "brownian", internal_nodes = TRUE)
image(Matrix(zapsmall(m1)))
dim(zapsmall(m1))
dim(m2)
```

If using all-nodes parameterization of phylogenetic precision matrix (to take advantage of sparsity/`gmrf()` in `RTMB`), we'd need to make sure that the 'species' factor included values (not included in data set) for the internal nodes ...

## To do

* replicate examples from `propto` paper, in RTMB (linear and functional effects of log(body mass) on running speed): additive model
* try out separable model via `dseparable`
* try out tensor-product model
* what happens if we include a fixed effect of clade?

```{r pic1}
ggplot(chdat, aes(runningspeed, bodymass, colour = clade)) +
  geom_point(size=4) +
  scale_x_log10() +
  scale_y_log10() +
  scale_colour_manual(values = c("darkred", "darkblue"))
```
