---
author: Ben Bolker
title: "How should we model combinations of phylogenetic correlations with other (non-scalar) random effects models?"
date: today
format:
    html:
       embed-resources: true
       code-fold: true
---

Tow (potential) parts to this:

* benchmarking alternative approaches to phylogenetic fits (Z-matrix, dense (tip-only) precision matrix, sparse (full tree) precision matrix ...
* alternative approaches to crossed (phylogenetic × something else): additive, separable, tensor-product

```{r pkgs, message=FALSE}
if (!require("MRFtools")) {
    stop('please remotes::install_github("eric-pedersen/MRFtools")')
    ##remotes::install_github("eric-pedersen/MRFtools")
}
if (!require("phyloglmm")) {
    stop('please remotes::install_github("wzmli/phyloglmm/pkg")')
}

## data / data manipulation
library(ade4)
library(ape)
library(dplyr)
## modeling
library(glmmTMB)
library(RTMB)
library(reformulas)
library(mgcv)
library(MRFtools)
library(Matrix)
## model diagnostics,
library(performance)
library(ggplot2); theme_set(theme_bw())
```

The McGillycuddy *et al.* manuscript on 'propto' random effects (i.e. random effects where the covariance matrix is known *a priori* up to a scalar multiplier) illustrates how to use `glmmTMB` + `propto` to fit models with a phylogenetic term.

```{r utils}
TMBfit <- function(obj, optimizer = nlminb, response = NULL) {
  fit <- optimizer(obj$par, obj$fn, obj$gr)
  ret <- list(fit = fit, obj = obj)
  if (!is.null(response)) attr(ret, "response") <- response
  class(ret) <- c("TMB", "list")
  ret
}
AIC.TMB <- function(x) { 2*x$fit$objective + 2*npar(x) }
logLik.TMB <- function(x) { -1*x$fit$objective }
fixef.TMB <- function(x) { x$obj$env$parList()$beta }
npar <- function(x, ...) {
  UseMethod("npar")
}
npar.TMB <- function(x) { length(x$fit$par) }
## these assume we have REPORT(mu) and REPORT(resid) and ADREPORT(mu)
predict.TMB <- function(x, se.fit = FALSE) {
  if (!se.fit) x$obj$report()$mu
  sdr <- TMB::sdreport(x$obj)
  sdr2 <- summary(sdr, select = "report")
  list(fit = sdr2[,"Estimate"],
       se.fit = sdr2[,"Std. Error"])
}  
```

```{r phyloZ_funs}
##' construct a Z matrix from a phylogeny
##' @name phylo_machinery
##' @param r a \code{phylo} object
##' @param stand standardize edge lengths by determinant of phylogenetic covariance matrix
##' @export
##' @importFrom Matrix t diag
##' @importFrom ape vcv
## FIXME:: rename to phylo_to_Z ?
phylo.to.Z <- function(r, stand = FALSE) {
  ntip <- length(r$tip.label)
  Zid <- Matrix::Matrix(0.0, ncol = length(r$edge.length), nrow = ntip)
  nodes <- (ntip + 1):max(r$edge)
  root <- nodes[!(nodes %in% r$edge[, 2])]
  for (i in 1:ntip) {
    cn <- i ## current node
    while (cn != root) {
      ce <- which(r$edge[, 2] == cn) ## find current edge
      Zid[i, ce] <- 1 ## set Zid to 1
      cn <- r$edge[ce, 1] ## find previous node
    }
  }
  tZid <- t(Zid)
  Z <- t(sqrt(r$edge.length) * tZid)
  if (stand) {
    V <- ape::vcv(r)
    ## V <- V/max(V)
    sig <- exp(as.numeric(determinant(V)["modulus"]) / ntip)
    ## sig <- det(V)^(1/ntip)
    ## all.equal(Z/sqrt(sig),
    ##  t(sqrt(r$edge.length / sig) * tZid), tolerance=2e-16)
    Z <- Z/sqrt(sig)
  }
  rownames(Z) <- r$tip.label
  colnames(Z) <- 1:length(r$edge.length)
  return(Z)
}
```

```{r setup}
gsub2 <- function(x, pattern, replacement) {
  gsub(pattern, replacement, x)
}

## typos/mismatches between data and tip labels
## (fix upstream?)
fixup_species <- function(x) {
  x |>
    gsub2("\\.", "_") |>
    gsub2("nelsoni$", "") |>
    gsub2("cinereorenteus", "cinereoargenteus") |>
    gsub2("hemonius", "hemionus") |>
    gsub2("thomsonii", "thompsonii") |>
    gsub2("Odicoileus", "Odocoileus") |>
    gsub2("_$", "")
}

data("carniherbi49", package = "ade4")
chtree <- read.tree(text=carniherbi49$tre2)
chtree$tip.label <- fixup_species(chtree$tip.label)
vcmat <- vcv(chtree)
chdat <- carniherbi49$tab2 |>
  mutate(log_bm = log(bodymass),
         log_rs = log(runningspeed),
         g = 1  ## dummy for grouping variable
         ) |>
  tibble::rownames_to_column("species") |>
  mutate(across(species, fixup_species)) |>
  mutate(across(species, \(x) factor(x, levels = rownames(vcmat))))
## check
stopifnot(identical(sort(rownames(vcmat)), sort(as.character(chdat$species))))
```

Model fit from McGillycuddy et al.:
```{r fit1}
fit_propto_glmmTMB <- glmmTMB(log_rs ~ log_bm + propto(0 + species | g, vcmat),
              data = chdat)
```

```{r fit1_check}
performance::check_model(fit_propto_glmmTMB,
                         check = c("linearity",
                                   "homogeneity", "pp_check",
                                   "qq"))
```

RTMB analogue:

```{r rtmb1}
X <- model.matrix(~ log_bm, data = chdat)
Z <- fac2sparse(chdat$species)
nllfun1 <- function(params) {
  getAll(params, chdat)
  mu <- drop(X %*% beta + Z %*% b)
  REPORT(mu)
  ADREPORT(mu)
  resid <- log_rs - mu
  REPORT(resid)
  pen <- -dmvnorm(b, 0, Sigma = vcmat, scale = exp(logpsd), log = TRUE)
  lik <- -sum(dnorm(log_rs, mean = mu, sd = exp(logsd), log = TRUE))
  lik + pen
}
p0 <- list(beta = rep(0, 2), b = rep(0, nrow(vcmat)), logsd=0, logpsd = 0)
obj <- MakeADFun(nllfun1, p0, silent = TRUE, random = "b")
fit_propto_TMB <- TMBfit(obj)
```

```{r glmmTMB_rtmb_check}
stopifnot(all.equal(fixef(fit_propto_TMB), fixef(fit_propto_glmmTMB)$cond,
                    check.attributes = FALSE, tolerance = 1e-7))
```

Thought I would try adding `clade` as a fixed effect, but it doesn't seem to
do much --- abandon that avenue for now.

## separable

```{r phylo_sep1}
us2 <- unstructured(2)
rt <- mkReTrms(findbars(~ (1 + log_bm | species)),
               fr = chdat, calc.lambdat = FALSE)
## scale=1 for identifiability (?)
f_phylo <- function(x) dmvnorm(x, 0, Sigma = vcmat, scale = 1, log = TRUE)
## need to pass params
mk_f_cov <- function(corval, logrsd) {
    function(x) dmvnorm(x, rep(0,2),
                        Sigma = us2$corr(corval), scale = exp(logrsd),
                        log = TRUE)
}
nllfun_sep <- function(params) {
  getAll(params, chdat)
  mu <- drop(X %*% beta + t(rt$Zt) %*% c(b))
  REPORT(mu)
  ADREPORT(mu)
  resid <- log_rs - mu
  ## REPORT(resid)
  pen <- -dseparable(f_phylo, mk_f_cov(corval, logrsd))(b)
  lik <- -sum(dnorm(log_rs, mean = mu, sd = exp(logsdres), log = TRUE))
  lik + pen
}
p2 <- list(beta = rep(0, 2), b = matrix(0,
                                        nrow = nrow(vcmat),
                                        ncol = 2),
           logrsd=rep(0, 2),
           logsdres = 0,
           corval=0)
obj2 <- MakeADFun(nllfun_sep, p2, silent = TRUE, random = "b")
fit_sep <- TMBfit(obj2)
```

Debugging `sdreport` stuff/methods ...
```{r}
sdr <- sdreport(fit_sep$obj)
```

```{r test_phyloZ}
f <- ~ 1 + (1+log_bm | species)
J <- eval(bquote(model.matrix(~.(findbars(f)[[1]][[2]]), data = chdat)))
pZ <- phylo.to.Z(chtree)
KR <- t(KhatriRao(t(pZ), t(J)))
image(KR)
```

## tensor-product

Not quite sure how I should do a tensor product between a phylo and a simple basis - do I need to drop the intercept?
Then what about the phylogenetic correlation in the intercepts?

Clark, Nicholas. “Phylogenetic Smoothing Using Mgcv.” GAMbler, 2024. https://ecogambler.netlify.app/blog/phylogenetic-smooths-mgcv/.


```{r mrf_penalty_test}
m1 <- mrf_penalty(chtree, "brownian", internal_nodes = FALSE)
m2 <- mrf_penalty(chtree, "brownian", internal_nodes = TRUE)
image(Matrix(zapsmall(m1)))
dim(zapsmall(m1))
dim(m2)
```

If using all-nodes parameterization of phylogenetic precision matrix (to take advantage of sparsity/`gmrf()` in `RTMB`), we'd need to make sure that the 'species' factor included values (not included in data set) for the internal nodes ...

## To do

* replicate examples from `propto` paper, in RTMB (linear and functional effects of log(body mass) on running speed): additive model
* try out separable model via `dseparable`
* try out tensor-product model
* what happens if we include a fixed effect of clade?

```{r pic1}
ggplot(chdat, aes(runningspeed, bodymass, colour = clade)) +
  geom_point(size=4) +
  scale_x_log10() +
  scale_y_log10() +
  scale_colour_manual(values = c("darkred", "darkblue"))
```
