---
title: "Wald vs likelihood ratio tests: examples"
author: "Ben Bolker"
date: today
format:
  html:
    embed-resources: true
    code-fold: true
knitr:
  opts_chunk:
    cache.comments: false    
bibliography: wald_lrt_examples.bib
nocite: |
  @hauckWaldTestApplied1977, @vaethUseWaldTest1985,
  @fearsReminderFallibilityWald1996, @pawitanReminderFallibilityWald2000,
  @yeeHauckDonnerEffect2022, @agrestiCategoricalDataAnalysis2013,
  @abanInferencesPowerAnalysis2009
---

Exploring (with Claude ...) the properties of Wald p-values (via `car::Anova()`) and likelihood ratio tests [LRT] (via `drop1()`): in particular, type-I error/coverage.

**tl;dr** there is surprisingly (to me) little difference in type-I error between Wald and LRT for sensible cases, although presumably we could find some if we tried hard enough. Coverage[^coverage] of confidence intervals for non-null cases shows bigger (but still surprisingly small) differences. The report explores that case for logistic regression, but the results are complicated. Logistic regression estimates are biased for small $n$; this
effect fights with the conservatism of Wald CIs due to Hauck-Donner effects [@hauckWaldTestApplied1977]. In the logistic case we also explore variant estimators that shrink the estimates toward zero (Firth/Jeffreys prior, $t$ prior), which mitigate upward small-sample bias of the MLE when the effects are moderate but lead to downward bias (by overcorrecting) when the effects are large ...

Even so, the advantages of the LRT/profile confidence intervals have more to do with power and interval shape than with type-I error/coverage. Over a sensible range of effect sizes for the binomial (say $0 \le \beta < 4$), with baseline probability $p_0 = 0.2$ and $n = 50$ or 100 per group, both methods get two-tailed coverage within about 0.02 of nominal, partly because some directional biases cancel out for the Wald CIs. (The patterns of one-sided coverage are more interesting/complicated.) With $n = 20$ per group, Wald is conservative, while the unpenalized LRT is anticonservative for $\beta$ up to about 4.5--5 (worst around $\beta = 3.5$, where the coverage drops to about 0.90) and conservative beyond that.

[^coverage]: the non-null analogue of type-I error, i.e. when the confidence interval includes fewer or more cases than the $100 (1-\alpha) \%$ it's supposed to

```{r setup}
#| message: false
library(glmmTMB)
library(car)
library(tidyverse)
library(futurize)
library(patchwork)

## avoid overcommitting threads
library(RhpcBLASctl)
blas_set_num_threads(1)
omp_set_num_threads(1)

## adjust this to something that makes sense on your machine (memory
## as well as cores), or set the WALD_WORKERS environment variable
nworkers <- as.integer(Sys.getenv("WALD_WORKERS",
                                  min(28, parallel::detectCores() - 1)))
plan(multicore, workers = nworkers)

theme_set(theme_bw())
zmargin <- theme(panel.spacing = grid::unit(0, "pt"))
## Okabe-Ito minus black and yellow
oi_cols <- c("#E69F00", "#56B4E9", "#009E73", "#0072B2", "#D55E00", "#CC79A7")
scale_colour_discrete <- function(...) scale_colour_manual(..., values = oi_cols)

## functions shared with wald_lrt_examples_lowp.R: the empty chunks
## below with matching labels get their code from here
knitr::read_chunk("wald_lrt_funs.R")
```

## Negative binomial GLMM: type I error under the null

Original example from Christoph Scherber: compare `drop1()`
(likelihood ratio test) and `car::Anova()` (Wald test) for the
species effect in a zero-inflated negative binomial GLMM.

```{r nb-example}
m2 <- glmmTMB(count ~ spp + mined + (1|site),
   zi=~spp + mined,
   family=nbinom2, data=Salamanders)
drop1(m2,test="Chisq")
Anova(m2)
```

```{r nb-funs}
```

Type I error for the full data set (n = `r nrow(Salamanders)`) and for
random subsamples of size 100:

```{r nb-sims-full}
#| cache: true
res <- full_sim()
```

```{r nb-sims-small, warning = FALSE}
#| cache: true
res_small <- full_sim(n = 100)
```

A few fits to the subsampled data have non-positive-definite Hessians,
giving NA p-values; these are dropped (`n_na` counts them).

```{r nb-results}
nb_sum <- function(r) {
  c(colMeans(r < 0.05, na.rm = TRUE), n_na = sum(!complete.cases(r)))
}
bind_rows(full = nb_sum(res),
          n100 = nb_sum(res_small),
          .id = "sample")
```
(The simulations take `r rpt_time(res)` and
`r rpt_time(res_small)` seconds, respectively, on 28 cores on my machine)

There is very little difference between the tests in this case, and at
n = 100 both are somewhat anticonservative. The Wald and LRT are
asymptotically equivalent under the null, so a null location effect
with a reasonable sample size/signal-to-noise ratio
is a friendly case for the Wald test; both
tests have slightly inflated type-I errors in the small sample case.

The bigger differences are in coverage and power for non-null cases.

## Binomial example

Two-group logistic regression (baseline probability 0.2, 20, 50 or 100
per group); the log-odds ratio [($\beta$)]{style="white-space: nowrap;"} ranges from 0 (null) to large
values, where the second group's probability approaches 1 and the
log-likelihood becomes strongly asymmetric (Hauck-Donner territory).
With a sparser baseline (e.g. 0.05) the LRT itself becomes
anticonservative under the null, muddying the comparison (see
[below](#sec-lowp) for a baseline of 0.01).

With $n$ observations per group there are only $(n+1)^2$ possible
outcomes $(y_0, y_1)$ (441, 2601 and 10201 for n = 20, 50 and 100), so we can compute the Wald and profile CIs for
every outcome once and then get the exact coverage for any true
$\beta$ by weighting each outcome by its binomial probability. This
removes the Monte Carlo error and allows a fine grid of $\beta$ values.

The two-group model is saturated, so the MLE, Wald SE and maximum
log-likelihood have closed forms; the profile likelihood only needs a
one-dimensional optimization over the intercept. Under separation
(any group with 0 or $n$ successes) the Wald CI is taken as
$(-\infty, \infty)$, the limit of the huge-SE intervals that
`glmmTMB` returns.

We also compute profile CIs from penalized likelihoods. Firth
regression maximizes the likelihood penalized by the Jeffreys prior
$|I(\theta)|^{1/2}$; here $|I| \propto p_0 q_0 p_1 q_1$, so the
Jeffreys prior factorizes into independent priors on the two group
logits $\eta_j$, each $\propto \sqrt{p_j q_j} = 1/(2 \cosh(\eta_j/2))$ (a
Beta(1/2, 1/2) prior on $p_j$, equivalent to adding 1/2 success and 1/2
failure to each group; on the logit scale this is a hyperbolic secant
distribution). `glmmTMB` doesn't offer this prior, but a $t$ prior
matched to its variance [($\pi^2$)]{style="white-space: nowrap;"} and kurtosis (7 df, scale
[$\pi \sqrt{5/7} \approx 2.66$)]{style="white-space: nowrap;"} is very close: the MAP estimate of a
single group's logit is within 0.05 (n = 20), 0.06 (n = 50) or 0.1 (n = 100) of Firth's
for every possible outcome. In `glmmTMB` the prior has to go on the
group logits, i.e.
`glmmTMB(y ~ 0 + g, family = binomial, priors = data.frame(prior = "t(0, 2.66, 7)", class = "fixef"))`;
independent priors on the intercept and log-odds ratio in the
`y ~ g` parameterization don't have the same structure.

For each penalty we compute both a profile CI and a penalized Wald CI
(MAP estimate $\pm 1.96$ SE, with the SE from the curvature of the
penalized log-likelihood). With the Firth penalty the information for
group $j$ is $(n+1)\tilde p_j \tilde q_j$, where
$\tilde p_j = (y_j + 1/2)/(n+1)$, so the penalized Wald CI is the
classic Haldane-Anscombe interval that adds 1/2 to every cell of the
$2 \times 2$ table.

```{r binom-plot-funs}
```

```{r binom-exact-funs}
#| cache: true
```

```{r binom-exact}
#| cache: true
#| dependson: binom-exact-funs
cis_2g <- map(c(20, 50, 100), exact_cis_n)
cov_2g <- map_dfr(cis_2g, exact_cov_n)
cis_2g <- bind_rows(cis_2g)
```

### Confidence intervals

The unpenalized estimates are biased upwards: the median MLE lies
slightly above the true value at every $\beta$, and it is infinite
whenever complete separation has probability above 1/2: at
$\beta = 5$ and 6 for n = 20 (probabilities 0.59 and 0.82) and at
$\beta = 6$ for n = 50 (0.61). The penalized (Firth and $t$ prior)
estimates are nearly identical to each other; they are close to the
true value for moderate $\beta$ but fall below it for large $\beta$,
where the penalty overcorrects. The intervals widen rapidly as $\beta$
increases, mostly upwards; where the median MLE is infinite, the median
unpenalized profile interval is unbounded above and the median
unpenalized Wald interval is $(-\infty, \infty)$. Within each penalty
the profile interval is asymmetric, extending further above the
estimate than below, while the Wald interval is symmetric, so the Wald
interval sits lower; in particular its upper limit is lower. All of
these patterns are stronger, and appear at smaller $\beta$, for smaller
n.

Median estimates and CI limits (over the exact sampling distribution)
for large true log-odds ratios; the black segments mark the true
values. Median limits of $\pm \infty$ are drawn as arrows running off
the edge of the panel, and median estimates of $+\infty$ (unpenalized:
$\beta = 5$ and 6 for n = 20, $\beta = 6$ for n = 50) aren't shown.

```{r binom-exact-medians}
#| fig-width: 9
#| out-width: 1036px
#| fig-column: page
med_2g <- map_dfr(3:6, exact_medians, cis = cis_2g)
plot_medians(med_2g)
```

### Coverage

One-sided coverage by tail: a well-calibrated 95% CI should have its
lower limit below (and its upper limit above) the true value
$\approx 97.5\%$ of the time. As a reference for how large a deviation
from 0.975 is noticeable, the gray band is the 95% (Gaussian) binomial
confidence interval around 0.975 for a simulation study with 2000
replicates.
Colours show the penalty (none, Firth/Jeffreys, or its $t$
approximation); solid lines are profile CIs and dashed lines are Wald
CIs. Comparing colours within a line type shows the effect of removing
the small-sample bias; comparing line types within a colour shows the
effect of the CI shape. The one-sided coverage axis is on a logit
scale, with values above 0.995 drawn at 0.995. The right-hand panels
show the two-sided coverage (the sum of the one-sided coverages minus
1), with the corresponding reference band around 0.95.

```{r binom-exact-plot}
#| fig-width: 10
#| fig-height: 7
#| out-width: 1151px
#| fig-column: page
## direct labels (by penalty) in the n = 50 lower-tail panel
pen_labs <- tibble(penalty = pen_levs,
                   label = c("unpenalized", "Firth", "t prior"),
                   beta = c(3, 1, 1), coverage = c(0.935, 0.992, 0.987),
                   n = 50)
plot_coverage(cov_2g, pen_labs)
```

### Low baseline probability {#sec-lowp}

The same calculations with a baseline probability of 0.01 (expected
0.2, 0.5 and 1 successes in the baseline group for n = 20, 50 and 100).
The CIs for each outcome don't depend on the baseline, so only the
weights change. (`wald_lrt_examples_lowp.R` does this as a
stand-alone script.) The unpenalized estimate is undefined (NaN) when
both groups have zero successes; these outcomes are dropped when
computing the medians, but they have probability at most 0.02
(n = 20, $\beta = 3$) at the values of $\beta$ shown.

```{r binom-lowp}
#| cache: true
#| dependson: binom-exact
p0_low <- 0.01
cov_lowp <- map_dfr(split(cis_2g, ~ n), exact_cov_n, p0 = p0_low)
med_lowp <- map_dfr(3:6, exact_medians, cis = cis_2g, p0 = p0_low)
```

The unpenalized median intervals are unbounded (in both directions
for Wald and above for the profile interval) at every $\beta$ shown
for n = 20 and 50; the penalized estimates fall below the true values.

```{r binom-lowp-medians}
#| fig-width: 9
#| out-width: 1036px
#| fig-column: page
plot_medians(med_lowp)
```

Coverage is strongly discrete (sawtoothed), and the lower tail is
conservative almost everywhere, so the two-sided coverage is mostly
determined by the upper tail. There, for $\beta \gtrsim 2$, the
penalized Wald intervals are anticonservative (two-sided coverage
about 0.92--0.93 for n = 50 and 100, lower for n = 20); the penalized
profile intervals are much closer to nominal, and the unpenalized
profile intervals are conservative. For n = 100 the unpenalized
profile interval is anticonservative for small $\beta$.

```{r binom-lowp-plot}
#| fig-width: 10
#| fig-height: 7
#| out-width: 1151px
#| fig-column: page
plot_coverage(cov_lowp)
```

## References

::: {#refs}
:::
