---
title: "CIs on ordinal prob predictions"
author: "Ben Bolker"
date: "`r format(Sys.time(),'%d %b %Y')`"
output:
  html_document:
    code_download: true
---

# Introduction

Constructing confidence intervals for the predicted probabilities of ordinal models turns out to be surprisingly (to me) hard. The basic problem is that the standard cumulative link models predict *cumulative* probabilities, not probabilities. It's easy to construct CIs for the cumulative probabilities by the standard methods: i.e. assuming MVN sampling distributions of parameters on the link scale, construct Normal CIs for the cumprob predictions and apply the inverse link function to get back to the [cumulative] probability scale). Unfortunately the probabilities themselves are differences of cumulative probabilities **on the probability scale** (not the link scale), so the standard tricks don't work.

People ask this question a lot (e.g. [here](https://stats.stackexchange.com/questions/201820/calculating-confidence-intervals-for-ordinal-logistic-regression-predictions), [here](https://stackoverflow.com/questions/53018272/confidence-intervals-of-predicted-probabilities-based-on-an-ordinal-regression-b))

The only simple choices I can think of for standard frequentist approaches are 

- some sort of delta-method approximation (these are dicey because they usually end up making a Normality assumption on the prediction scale, which can be bad: it's conceivable that one exists that's better than the one the `emmeans` package uses)
- a pseudo "prediction interval" (*sensu* Lande et al.) based on sampling random MVN values of the parameter vector
- Jonathan Dushoff points out that if you can be satisfied with effects plots that show the change in probability from a specified baseline and incorporate the uncertainty of only one predictor, this can be done in the classical framework.

The other choice is to use a Bayesian method, which is illustrated below.

# Model fitting

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r pkgs,message=FALSE}
library(ordinal)
library(brms)
library(MASS)  ## polr: *before* dplyr
library(dplyr)
library(tidyr)
library(tidybayes)
library(rstanarm)
library(broom)
library(broom.mixed)
library(dotwhisker)
library(ggplot2); theme_set(theme_bw())
```

```{r echo=FALSE,eval=FALSE}
## load cache in interactive session
lazyLoad('ordinal_probpredCI_cache/html/fits_e463d7f40e8207120a9c109e5dabe5ef')
```

Fit the same model in `ordinal` and `brms`
using the basic `wine` example from the `ordinal` package:

```{r fits, message=FALSE,cache=TRUE}
m_clm <- clm(rating ~ temp * contact, data = wine)
m_brm <- brm(rating ~ temp * contact,
          family =cumulative(threshold="flexible"),
          data = wine,
          refresh=0   ## quiet
          )
```

Add `MASS::polr` and `rstanarm::stan_polr` for comparison:

```{r other_fits,cache=TRUE,message=FALSE,warning=FALSE}
c_MASS <- polr(rating ~ temp * contact, data = wine)
c_rstan <- stan_polr(rating ~ temp * contact, data = wine,
                     prior=R2(0.2,"mean"),
                     refresh=0)
```

Could also try this with `MCMCglmm`.

Trying to do a dot-whisker plot to compare parameters (but some technical problems ...)
```{r FL}
fitlist <- list(clm=m_clm,brm=m_brm,MASS=c_MASS,
                rstan=c_rstan)
## FIXME: tidy() has problems with m_brm (bad 'term' var)
## FIXME: dwplot() has problems with m_clm
dwplot(fitlist[3:4])
```

Hack our own dot-whisker plot (don't need `ggstance::geom_pointrangeh`
because we're not trying to facet, you we can `coord_flip()` instead)
```{r dw2,warning=FALSE}
tdf <- purrr::map_dfr(fitlist,tidy,.id="model",conf.int=TRUE)
## hack to fix brm term names
tdf[tdf$model=="brm","term"] <- tdf[tdf$model=="clm","term"]
pd <- position_dodge(width=0.4)
(ggplot(tdf,aes(x=term,y=estimate,colour=model))
   ## redundant (some models missing intercept CIs)
  + geom_point(position=pd,size=3)
  + geom_pointrange(aes(ymin=conf.low,ymax=conf.high),
                    position=position_dodge(width=0.4))
  + coord_flip()
  + scale_colour_brewer(palette="Dark2")
  + labs(x="")
)
```

*Slightly* different predictions (median vs MLE; priors in rstan and brms).

## predictions with CIs

Prediction data frame (for marginal values):

```{r pframe}
pframe <- with(wine,expand.grid(temp=levels(temp),
                                contact=levels(contact)))
```

Reorg/gather predictions from both approaches:

```{r reorg}
pred_clm <- (data.frame(pframe,
                  setNames(as.data.frame(predict(m_clm,newdata=pframe)),1:5),
                  check.names=FALSE)
    %>% tidyr::gather(rating,est,-c(temp,contact), factor_key=TRUE)
)

pred_brm <- (tidybayes::fitted_draws(m_brm,pframe,category="rating")
    %>% select(temp,contact,rating,.value)
    %>% group_by(temp,contact,rating)
    %>% summarise(lo=quantile(.value,0.025),
                  est=median(.value),
                  hi=quantile(.value,0.975))
    %>% ungroup
)
pred_all <- dplyr::bind_rows(list(clm=pred_clm,brm=pred_brm), .id="model")
```

Draw the picture:

```{r pix}
gg0 <- (ggplot(pred_all,aes(temp,est,colour=rating))
    + geom_point(aes(shape=model))
    + geom_line(aes(group=interaction(model,rating),linetype=model))
 ## NAs in clm predictions mess this up, so restrict to pred_brm
    + geom_ribbon(data=pred_brm,aes(ymin=lo,ymax=hi,
                                    fill=rating,group=rating),
                  colour=NA,alpha=0.2)
)
gg0 + facet_wrap(~contact, labeller=label_both)
gg0 + facet_grid(contact ~ rating,
                 labeller=label_both) + theme(panel.spacing=grid::unit(0,"lines"))
```

# to do

- submit GH issues/fixes for glitches in  `broom.mixed` and `dwplot`
- try with `MCMCglmm`?
- try out various methods for frequentist fits
- compare `emmeans` results (delta method)
