---
title: Power-logistic regression link
author: Ben Bolker
bibliography: nestsurv.bib
---

R code for a link function of the form 
$$
\mu = (1+\exp(-\eta))^{-\theta}
$$
(i.e., a logistic function raised to the power $\theta$, where
$\theta$ is a measure of the exposure).
This code was
originally adapted from Mark Herzog's code at `http://www.npwrc.usgs.gov/resource/birds/nestsurv/download/CreateLogisticExposureFamily.R`, which
in turn translated Shaffer's SAS code,
but this page is now only available via the
[Wayback machine](https://web.archive.org/web/*/http://www.npwrc.usgs.gov/resource/birds/nestsurv/download/CreateLogisticExposureFamily.R)

Shaffer's original paper [@shaffer_unified_2004] references
[@burhans_habitat_1999] for data on yellow-breasted chat nest
survival, and references http://www.npwrc.usgs.gov/resource/tools/nestsurv/nestsurv.htm ; I managed to retrieve the data and SAS code via the
Wayback machine.

```{r logexp_fun}
##' @param exposure (vector of) exposure times
##' @param fast (logical): use R-coded functions for stability,
##' or call internal C code for speed?
logexp <- function(exposure = 1, fast=FALSE)
{
    linkfun <- function(mu) qlogis(mu^(1/exposure))
    ## FIXME: is there some trick we can play here to allow
    ##   evaluation in the context of the 'data' argument?
    linkinv <- function(eta)  plogis(eta)^exposure
    logit_mu_eta <- function(eta) {
        if (FALSE) {
            ifelse(abs(eta)>30,.Machine$double.eps,
                   exp(eta)/(1+exp(eta))^2)
        } else {
            .Call(stats:::C_logit_mu_eta, eta, PACKAGE = "stats")
        }
    }
    mu.eta <- function(eta) {       
        exposure * plogis(eta)^(exposure-1) *
            logit_mu_eta(eta)
    }
    valideta <- function(eta) TRUE
    link <- paste("logexp(", deparse(substitute(exposure)), ")",
                   sep="")
    structure(list(linkfun = linkfun, linkinv = linkinv,
                   mu.eta = mu.eta, valideta = valideta, 
                   name = link),
              class = "link-glm")
}
```

Here is an example using the chat data from Shaffer (2004),
but these data are now unavailable ...

```{r eval=FALSE}
nestdata<-read.table("http://data.prbo.org/tools/NestSurvival/chat.txt")
chat.glm.logexp <- glm(survive/trials~parastat+nest_ht*patsize,
                       family=binomial(logexp(days=nestdata$expos)),
                       data=nestdata)
# if you have MASS installed
library("MASS")
chat.step<-stepAIC(chat.glm.logexp,
                   scope=list(upper=~parastat+nest_ht*patsize,lower=~1))
chat.step$anova
summary(chat.step)
```

SAS results:

```
              Criteria For Assessing Goodness Of Fit

    Criterion                 DF           Value        Value/DF
    Deviance                 289        193.9987          0.6713
    Scaled Deviance          289        193.9987          0.6713
    Pearson Chi-Square       289        537.8609          1.8611
    Scaled Pearson X2        289        537.8609          1.8611
    Log Likelihood                      -96.9994
    Algorithm converged.
                       Analysis Of Parameter Estimates

                                Standard Wald 95%   Chi-
Parameter   DF  Estimate Error        Limits       Square Pr > ChiSq
#Intercept   1  2.6973   0.2769   2.1546  3.2399   94.92  <.0001
parastat0    1 -1.0350   0.5201  -2.0544  -0.0155  3.96   0.0466
parastat1    0  0.0000   0.0000   0.0000  0.0000     .      .
patsizelarge 1  1.0844   0.5094   0.0861  2.0827   4.53   0.0333
patsizesmall 0  0.0000   0.0000   0.0000  0.0000     .       .
Scale        0  1.0000   0.0000   1.0000  1.0000

#NOTE: The scale parameter was held fixed.
```

## References
